---
title: "Accelerated Protein Evolution — High dN/dS and Branch-Specific Positive Selection"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    fig_caption: true
params:
  # Named list: species_label → output_dir from render_analysis.R
  # All input files (branch_specific_MKT_results.tsv, candidates_*.tsv) are
  # read automatically from these directories.
  species_output_dirs: !r list()
  # z-score threshold above which a gene is classified as "high dN/dS"
  z_threshold: 1.5
  # p-value threshold for the branch model likelihood ratio test (LRT)
  lrt_p_threshold: 0.05
  # Directory where output tables are written
  output_dir: "results/high_dn_ds"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 9, fig.height = 6)
```

```{r libraries}
library(tidyverse)
library(ComplexUpset)
library(kableExtra)
library(patchwork)
```

This notebook quantifies the overlap between branch-specific positive selection
signals (from McDonald-Kreitman tests) and protein sequence evolution
acceleration signals (high branch-specific dN/dS, branch model LRT).

**Analyses performed:**

1. Per-species high dN/dS identification (z-score on log-scale)
2. Per-species enrichment of MKT candidates in the high dN/dS category
3. Per-species branch model likelihood ratio test (LRT) for acceleration
4. Cross-species lineage-specific dN/dS patterns (when ≥2 species)
5. Cross-species UpSet plots combining MKT methods and high dN/dS

---

## 1  Load data

```{r params_check}
if (length(params$species_output_dirs) == 0)
  stop("params$species_output_dirs is empty. Edit render_high_dn_ds.R and provide ",
       "at least one species.")

n_species <- length(params$species_output_dirs)
sp_labels <- names(params$species_output_dirs)

cat(sprintf("Species to analyse: %d\n", n_species))
for (sp in sp_labels)
  cat(sprintf("  • %s → %s\n", sp, params$species_output_dirs[[sp]]))

cat(sprintf("\nParameters:\n"))
cat(sprintf("  z_threshold      = %.2f\n", params$z_threshold))
cat(sprintf("  lrt_p_threshold  = %.3f\n", params$lrt_p_threshold))
```

```{r load_species_data}
# Build file paths from output_dir, then load all species data into named lists.
# File names match the standardised output of branch_specific_MKT_analysis.Rmd.

load_species_data <- function(sp, out_dir) {
  files <- list(
    stats    = file.path(out_dir, "branch_specific_MKT_results.tsv"),
    cand_all = file.path(out_dir, "candidates_all.tsv"),
    cand_bs  = file.path(out_dir, "candidates_branch_specific.tsv")
  )
  missing <- names(Filter(Negate(file.exists), files))
  if (length(missing) > 0)
    stop("Files not found for '", sp, "': ", paste(missing, collapse = ", "),
         "\nCheck that render_analysis.R was run before this script.")

  list(
    stats    = read_tsv(files$stats,    col_types = cols(.default = col_double(),
                                                          gene_id  = col_character(),
                                                          ortholog = col_character()),
                        na = c("NA", "NaN", "")),
    cand_all = read_tsv(files$cand_all, col_types = cols(.default = col_character()),
                        na = c("NA", "NaN", "")),
    cand_bs  = read_tsv(files$cand_bs,  col_types = cols(.default = col_character()),
                        na = c("NA", "NaN", ""))
  )
}

species_data <- setNames(
  mapply(load_species_data, sp_labels, params$species_output_dirs,
         SIMPLIFY = FALSE),
  sp_labels
)

for (sp in sp_labels) {
  cat(sprintf("\n[%s]\n", sp))
  cat(sprintf("  Genes in stats table : %d\n", nrow(species_data[[sp]]$stats)))
  cat(sprintf("  All candidates       : %d\n", nrow(species_data[[sp]]$cand_all)))
  cat(sprintf("  BS candidates        : %d\n", nrow(species_data[[sp]]$cand_bs)))
}
```

---

## 2  Analyzable gene sets

Genes are **analyzable** for a given test when they pass the McDonald-Kreitman
filter:
- Pn ≥ 1 AND Ps ≥ 1 AND Dn ≥ 1 AND Ds ≥ 1 AND Pn + Ps ≥ 5

Four filter variants are computed using the Pn/Dn/Ds sources for each test:

| Set | Pn | Dn / Ds |
|-----|-----|---------|
| `bs_mkt` | Pn (raw) | Dn\_branch\_specific / Ds\_branch\_specific |
| `bs_imp_mkt` | PnNeutral | Dn\_branch\_specific / Ds\_branch\_specific |
| `std_mkt` | Pn (raw) | dn\_counts / ds\_counts |
| `std_imp_mkt` | PnNeutral | dn\_counts / ds\_counts |

```{r analyzable_sets}
mk_filter <- function(Pn, Ps, Dn, Ds) {
  !is.na(Pn) & Pn >= 1 & !is.na(Ps) & Ps >= 1 &
  !is.na(Dn) & Dn >= 1 & !is.na(Ds) & Ds >= 1 &
  (Pn + Ps) >= 5
}

get_analyzable <- function(stats_df) {
  list(
    bs_mkt = stats_df %>%
      filter(mk_filter(num_pol_NS, num_pol_S,
                       Dn_branch_specific, Ds_branch_specific)) %>%
      pull(gene_id),
    bs_imp_mkt = stats_df %>%
      filter(mk_filter(PnNeutral, num_pol_S,
                       Dn_branch_specific, Ds_branch_specific)) %>%
      pull(gene_id),
    std_mkt = stats_df %>%
      filter(mk_filter(num_pol_NS, num_pol_S, dn_counts, ds_counts)) %>%
      pull(gene_id),
    std_imp_mkt = stats_df %>%
      filter(mk_filter(PnNeutral, num_pol_S, dn_counts, ds_counts)) %>%
      pull(gene_id)
  )
}

analyzable <- lapply(sp_labels, function(sp)
  get_analyzable(species_data[[sp]]$stats)
)
names(analyzable) <- sp_labels

# Summary table
az_tbl <- bind_rows(lapply(sp_labels, function(sp) {
  tibble(
    Species         = sp,
    bs_mkt          = length(analyzable[[sp]]$bs_mkt),
    bs_imp_mkt      = length(analyzable[[sp]]$bs_imp_mkt),
    std_mkt         = length(analyzable[[sp]]$std_mkt),
    std_imp_mkt     = length(analyzable[[sp]]$std_imp_mkt),
    total_genes     = nrow(species_data[[sp]]$stats)
  )
}))

az_tbl %>%
  kbl(caption = "Number of genes passing each MKT filter") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 3  Per-species high dN/dS analysis

### 3.1  dN/dS distributions and z-scores

For each species, the high dN/dS universe consists of genes analyzable in at
least one branch-specific test (`bs_mkt` OR `bs_imp_mkt`) that have a valid
`DnDs_branch_specific_counts` value. Z-scores are computed on the
log₁₊ₓ-transformed values.

```{r compute_dnds_universe}
compute_dnds_universe <- function(sp) {
  az_bs  <- union(analyzable[[sp]]$bs_mkt, analyzable[[sp]]$bs_imp_mkt)
  stats  <- species_data[[sp]]$stats

  stats %>%
    filter(gene_id %in% az_bs, !is.na(DnDs_branch_specific_counts)) %>%
    mutate(
      log_dNdS  = log1p(DnDs_branch_specific_counts),
      z_dNdS    = as.numeric(scale(log_dNdS)),
      high_dNdS = z_dNdS > params$z_threshold
    )
}

dnds_universe <- lapply(sp_labels, compute_dnds_universe)
names(dnds_universe) <- sp_labels

for (sp in sp_labels) {
  u <- dnds_universe[[sp]]
  cat(sprintf("[%s] dN/dS universe: %d genes  |  high dN/dS (z > %.1f): %d (%.1f%%)\n",
              sp, nrow(u), params$z_threshold,
              sum(u$high_dNdS),
              100 * mean(u$high_dNdS)))
}
```

```{r plot_dnds_distributions, fig.cap="Branch-specific dN/dS distributions. Red dashed line: z-score threshold. Candidates (branch-specific positive selection) are highlighted."}
plots_dist <- lapply(sp_labels, function(sp) {
  u <- dnds_universe[[sp]]
  bs_genes <- species_data[[sp]]$cand_bs$gene_id

  u %>%
    mutate(candidate = gene_id %in% bs_genes) %>%
    ggplot(aes(x = log_dNdS, fill = candidate, colour = candidate)) +
    geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
    geom_vline(
      xintercept = log1p(quantile(u$DnDs_branch_specific_counts,
                                   pnorm(params$z_threshold), na.rm = TRUE)),
      linetype = "dashed", colour = "red"
    ) +
    scale_fill_manual(
      values  = c("TRUE" = "#C44E52", "FALSE" = "#4C72B0"),
      labels  = c("TRUE" = "BS candidate", "FALSE" = "Background")
    ) +
    scale_colour_manual(
      values  = c("TRUE" = "#C44E52", "FALSE" = "#4C72B0"),
      guide   = "none"
    ) +
    labs(title   = sp,
         x       = expression(log[1+x](dN/dS[branch])),
         y       = "Genes",
         fill    = NULL) +
    theme_classic(base_size = 11) +
    theme(legend.position = "bottom")
})

wrap_plots(plots_dist, ncol = min(2, n_species))
```

### 3.2  Candidate enrichment in high dN/dS

**Fisher's exact test** (one-sided greater): are branch-specific MKT candidates
enriched among genes with high dN/dS?

**Wilcoxon rank-sum test**: do candidates have higher dN/dS than background?

```{r enrichment_high_dnds, results='asis'}
dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)

enrichment_results <- lapply(sp_labels, function(sp) {
  u          <- dnds_universe[[sp]]
  bs_genes   <- intersect(species_data[[sp]]$cand_bs$gene_id, u$gene_id)
  high_genes <- u$gene_id[u$high_dNdS]
  bg_genes   <- u$gene_id

  non_cand   <- setdiff(bg_genes, bs_genes)

  # 2×2 contingency table
  a <- sum(bs_genes  %in% high_genes)
  b <- sum(bs_genes  %in% setdiff(bg_genes, high_genes))
  c <- sum(non_cand  %in% high_genes)
  d <- sum(non_cand  %in% setdiff(bg_genes, high_genes))
  mat <- matrix(c(a, b, c, d), nrow = 2, byrow = TRUE,
                dimnames = list(c("BS candidate", "Background"),
                                c("High dN/dS",  "Not high dN/dS")))

  fisher <- fisher.test(mat, alternative = "greater")
  wilcox <- wilcox.test(DnDs_branch_specific_counts ~ (gene_id %in% bs_genes),
                        data = u)

  medians <- u %>%
    mutate(group = if_else(gene_id %in% bs_genes, "BS candidate", "Background")) %>%
    group_by(group) %>%
    summarise(n = n(), median_dNdS = round(median(DnDs_branch_specific_counts), 4))

  list(
    species   = sp,
    n_high    = sum(u$high_dNdS),
    n_cand_bs = length(bs_genes),
    n_overlap = a,
    mat       = mat,
    fisher    = fisher,
    wilcox    = wilcox,
    medians   = medians
  )
})
names(enrichment_results) <- sp_labels

# Print results per species
for (sp in sp_labels) {
  r <- enrichment_results[[sp]]
  cat(sprintf("\n### %s\n\n", sp))
  cat(sprintf("High dN/dS genes (z > %.1f): **%d**  |  ",
              params$z_threshold, r$n_high))
  cat(sprintf("BS candidates: **%d**  |  Overlap: **%d**\n\n",
              r$n_cand_bs, r$n_overlap))
  print(kbl(r$mat) %>% kable_styling(bootstrap_options = c("striped", "condensed")))
  cat(sprintf("\nFisher OR = %.3f  p = %.4f\n",
              r$fisher$estimate, r$fisher$p.value))
  cat(sprintf("Wilcoxon p = %.4f\n\n", r$wilcox$p.value))
  print(kbl(r$medians, caption = "Median dN/dS") %>%
          kable_styling(bootstrap_options = c("striped", "condensed")))
}

# Write combined enrichment summary
enrich_tbl <- bind_rows(lapply(sp_labels, function(sp) {
  r <- enrichment_results[[sp]]
  tibble(
    species       = sp,
    n_universe    = nrow(dnds_universe[[sp]]),
    n_high_dNdS   = r$n_high,
    n_candidates  = r$n_cand_bs,
    n_overlap     = r$n_overlap,
    fisher_OR     = round(r$fisher$estimate, 4),
    fisher_pvalue = r$fisher$p.value,
    wilcox_pvalue = r$wilcox$p.value
  )
}))

write_tsv(enrich_tbl, file.path(params$output_dir, "enrichment_high_dNdS.tsv"), na = "NA")
```

---

## 4  Branch model LRT (acceleration vs deceleration)

Likelihood ratio test comparing the branch model (lineage-specific dN/dS)
against the M0 model (global dN/dS):

$$LRT = 2 \times (lnL_{branch} - lnL_{M0}) \sim \chi^2_{df=1}$$

A gene is **accelerated** when *p* < `r params$lrt_p_threshold` AND
omega\_branch > omega\_M0.

```{r branch_model_lrt}
compute_lrt <- function(sp) {
  stats <- species_data[[sp]]$stats

  stats %>%
    filter(!is.na(lnL_branch_model), !is.na(lnL_m0_model),
           !is.na(omega),            !is.na(omega_m0)) %>%
    mutate(
      lrt          = 2 * (lnL_branch_model - lnL_m0_model),
      lrt_pvalue   = pchisq(lrt, df = 1, lower.tail = FALSE),
      significant  = lrt_pvalue < params$lrt_p_threshold,
      acceleration = omega > omega_m0,
      lrt_category = case_when(
        significant & acceleration  ~ "Accelerated",
        significant & !acceleration ~ "Decelerated",
        TRUE                        ~ "Neutral"
      )
    )
}

lrt_data <- lapply(sp_labels, compute_lrt)
names(lrt_data) <- sp_labels

# Summary per species
lrt_summary <- bind_rows(lapply(sp_labels, function(sp) {
  lrt_data[[sp]] %>%
    count(lrt_category, name = "n") %>%
    mutate(
      species = sp,
      pct     = round(100 * n / sum(n), 1)
    ) %>%
    select(species, lrt_category, n, pct)
}))

lrt_summary %>%
  kbl(caption = sprintf("Branch model LRT — classification (p < %.3f)",
                        params$lrt_p_threshold)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r plot_lrt_omega, fig.cap="omega (branch model, focal species) distributions by LRT category."}
omega_plots <- lapply(sp_labels, function(sp) {
  lrt_data[[sp]] %>%
    filter(!is.na(omega), omega < 10) %>%      # cap extreme outliers for display
    ggplot(aes(x = omega, fill = lrt_category, colour = lrt_category)) +
    geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
    geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
    scale_fill_manual(values = c(
      Accelerated = "#C44E52", Decelerated = "#4C72B0", Neutral = "#8C8C8C"
    )) +
    scale_colour_manual(values = c(
      Accelerated = "#C44E52", Decelerated = "#4C72B0", Neutral = "#8C8C8C"
    ), guide = "none") +
    facet_wrap(~lrt_category, scales = "free_y") +
    labs(title = sp, x = "omega (focal branch)", y = "Genes", fill = NULL) +
    theme_classic(base_size = 11) +
    theme(legend.position = "none")
})

wrap_plots(omega_plots, ncol = 1)
```

### 4.1  Candidates enriched in accelerated genes

```{r lrt_enrichment, results='asis'}
lrt_enrichment <- lapply(sp_labels, function(sp) {
  lrt_df   <- lrt_data[[sp]]
  bs_genes <- intersect(species_data[[sp]]$cand_bs$gene_id, lrt_df$gene_id)
  accel    <- lrt_df$gene_id[lrt_df$lrt_category == "Accelerated"]

  non_cand <- setdiff(lrt_df$gene_id, bs_genes)
  a <- sum(bs_genes %in% accel)
  b <- length(bs_genes) - a
  c <- sum(non_cand  %in% accel)
  d <- length(non_cand)  - c
  mat <- matrix(c(a, b, c, d), nrow = 2, byrow = TRUE,
                dimnames = list(c("BS candidate", "Background"),
                                c("Accelerated", "Not accelerated")))

  list(mat = mat, fisher = fisher.test(mat, alternative = "greater"))
})
names(lrt_enrichment) <- sp_labels

for (sp in sp_labels) {
  r <- lrt_enrichment[[sp]]
  cat(sprintf("\n### %s (LRT enrichment)\n\n", sp))
  print(kbl(r$mat) %>% kable_styling(bootstrap_options = c("striped", "condensed")))
  cat(sprintf("\nFisher OR = %.3f  p = %.4f\n\n", r$fisher$estimate, r$fisher$p.value))
}

# Write LRT classifications and summary
lrt_tbl <- bind_rows(lapply(sp_labels, function(sp) {
  lrt_data[[sp]] %>%
    select(gene_id, ortholog, omega, omega_m0, omega_diff,
           lnL_branch_model, lnL_m0_model, lrt, lrt_pvalue, lrt_category) %>%
    mutate(species = sp)
}))

write_tsv(lrt_tbl, file.path(params$output_dir, "lrt_classifications.tsv"), na = "NA")

lrt_enrich_tbl <- bind_rows(lapply(sp_labels, function(sp) {
  r <- lrt_enrichment[[sp]]
  tibble(
    species       = sp,
    n_accelerated = r$mat["BS candidate", "Accelerated"] + r$mat["Background", "Accelerated"],
    n_candidates  = sum(r$mat["BS candidate", ]),
    n_overlap     = r$mat["BS candidate", "Accelerated"],
    fisher_OR     = round(r$fisher$estimate, 4),
    fisher_pvalue = r$fisher$p.value
  )
}))

write_tsv(lrt_enrich_tbl, file.path(params$output_dir, "enrichment_lrt_acceleration.tsv"), na = "NA")
```

### 4.2  Triple overlap: high dN/dS ∩ accelerated ∩ BS candidates

```{r triple_overlap, results='asis'}
for (sp in sp_labels) {
  high_genes <- dnds_universe[[sp]]$gene_id[dnds_universe[[sp]]$high_dNdS]
  accel      <- lrt_data[[sp]]$gene_id[lrt_data[[sp]]$lrt_category == "Accelerated"]
  bs_cands   <- species_data[[sp]]$cand_bs$gene_id

  triple <- Reduce(intersect, list(high_genes, accel, bs_cands))

  cat(sprintf("\n### %s\n\n", sp))
  cat(sprintf("High dN/dS: %d  |  Accelerated: %d  |  BS candidates: %d\n",
              length(high_genes), length(accel), length(bs_cands)))
  cat(sprintf("High dN/dS ∩ Accelerated       : %d\n", length(intersect(high_genes, accel))))
  cat(sprintf("High dN/dS ∩ BS candidates     : %d\n", length(intersect(high_genes, bs_cands))))
  cat(sprintf("Accelerated ∩ BS candidates    : %d\n", length(intersect(accel, bs_cands))))
  cat(sprintf("**Triple overlap**             : **%d**\n\n", length(triple)))

  if (length(triple) > 0) {
    trip_df <- species_data[[sp]]$stats %>%
      filter(gene_id %in% triple) %>%
      select(gene_id, ortholog, DnDs_branch_specific_counts, omega, lrt_pvalue = any_of("lrt_pvalue")) %>%
      left_join(select(lrt_data[[sp]], gene_id, lrt_pvalue), by = "gene_id")
    cat("Genes in triple overlap:\n\n")
    print(kbl(trip_df) %>% kable_styling(bootstrap_options = c("striped", "condensed")))
  }
}
```

---

```{r cross_species_check, include=FALSE}
# Set flag used in all subsequent sections
run_cross_species <- n_species >= 2
```

## 5  Cross-species analysis

```{r cross_species_skip, results='asis', eval=!run_cross_species}
cat("> *Cross-species analysis requires ≥2 species in `params$species_output_dirs`.",
    "Currently only 1 species is provided — this section is skipped.*")
```

### 5.1  Ortholog pairs and lineage-specific dN/dS patterns

For each pair of focal species, ortholog pairs are identified using the
`{other_species}_ortholog` columns present in each stats table. The universe
is restricted to gene pairs analyzable in the branch-specific test in BOTH
species.

```{r cross_species_pairs, eval=run_cross_species}
# For each ordered pair (A, B), derive the column name pointing from A to B.
# The stats table of species A contains a column "{short(B)}_ortholog"
# where short(B) = last word of B (e.g. "mutica" from "Aegilopsmutica").
short_name <- function(sp) tolower(sub(".*_", "", sp))

# Find the ortholog-column name in sp_A that points to sp_B genes
find_ortho_col <- function(sp_a, stats_a, sp_b) {
  candidates <- paste0(short_name(sp_b), "_ortholog")
  found <- intersect(candidates, colnames(stats_a))
  if (length(found) == 0)
    warning("No ortholog column for '", sp_b, "' found in '", sp_a, "' stats table.")
  found[1]
}

# Build dN/dS universe with z-scores for each species
dnds_z <- lapply(sp_labels, function(sp) {
  dnds_universe[[sp]] %>%
    transmute(gene_id, z_dNdS, high_dNdS, dNdS = DnDs_branch_specific_counts)
})
names(dnds_z) <- sp_labels

# Build ortholog pairs for all unique pairs of focal species
all_pairs <- combn(sp_labels, 2, simplify = FALSE)

pair_dfs <- lapply(all_pairs, function(pair) {
  sp_a <- pair[1]; sp_b <- pair[2]
  stats_a <- species_data[[sp_a]]$stats

  ortho_col <- find_ortho_col(sp_a, stats_a, sp_b)
  if (is.na(ortho_col)) return(NULL)

  pairs <- stats_a %>%
    filter(!is.na(.data[[ortho_col]])) %>%
    transmute(
      gene_a    = gene_id,
      gene_b    = .data[[ortho_col]],
      ortholog  = ortholog
    ) %>%
    filter(gene_a %in% dnds_z[[sp_a]]$gene_id,
           gene_b %in% dnds_z[[sp_b]]$gene_id) %>%
    distinct()

  pairs %>%
    left_join(rename(dnds_z[[sp_a]], z_a = z_dNdS, high_a = high_dNdS, dNdS_a = dNdS),
              by = c("gene_a" = "gene_id")) %>%
    left_join(rename(dnds_z[[sp_b]], z_b = z_dNdS, high_b = high_dNdS, dNdS_b = dNdS),
              by = c("gene_b" = "gene_id")) %>%
    filter(!is.na(z_a), !is.na(z_b)) %>%
    mutate(
      a_specific = z_a >  params$z_threshold & z_b < -params$z_threshold,
      b_specific = z_b >  params$z_threshold & z_a < -params$z_threshold,
      shared_high = z_a > params$z_threshold & z_b >  params$z_threshold,
      shared_low  = z_a < -params$z_threshold & z_b < -params$z_threshold,
      sp_a = sp_a,
      sp_b = sp_b
    )
})
names(pair_dfs) <- sapply(all_pairs, paste, collapse = "_vs_")
pair_dfs <- Filter(Negate(is.null), pair_dfs)
```

```{r cross_species_summary_table, eval=run_cross_species, results='asis'}
for (pair_name in names(pair_dfs)) {
  df <- pair_dfs[[pair_name]]
  sp_a <- df$sp_a[1]; sp_b <- df$sp_b[1]
  cat(sprintf("\n#### %s vs %s\n\n", sp_a, sp_b))
  cat(sprintf("Ortholog pairs analysable in both species: **%d**\n\n", nrow(df)))

  tbl <- tibble(
    Pattern           = c(sprintf("%s-specific high dN/dS", sp_a),
                           sprintf("%s-specific high dN/dS", sp_b),
                           "Shared high dN/dS",
                           "Shared low dN/dS",
                           "Neither"),
    n                 = c(sum(df$a_specific), sum(df$b_specific),
                           sum(df$shared_high), sum(df$shared_low),
                           sum(!df$a_specific & !df$b_specific &
                               !df$shared_high & !df$shared_low)),
    pct               = round(100 * n / nrow(df), 1)
  )
  print(kbl(tbl, caption = paste(sp_a, "vs", sp_b)) %>%
          kable_styling(bootstrap_options = c("striped", "condensed")))
}
```

```{r plot_cross_dnds_scatter, eval=run_cross_species, fig.cap="Cross-species dN/dS z-score scatter. Red: lineage-specific high dN/dS; grey: other orthologs."}
scatter_plots <- lapply(names(pair_dfs), function(pair_name) {
  df   <- pair_dfs[[pair_name]]
  sp_a <- df$sp_a[1]; sp_b <- df$sp_b[1]

  df %>%
    mutate(pattern = case_when(
      a_specific  ~ paste(sp_a, "specific"),
      b_specific  ~ paste(sp_b, "specific"),
      shared_high ~ "Shared high",
      shared_low  ~ "Shared low",
      TRUE        ~ "Neutral"
    )) %>%
    ggplot(aes(x = z_a, y = z_b, colour = pattern)) +
    geom_point(alpha = 0.4, size = 0.9) +
    geom_hline(yintercept = c(-params$z_threshold, params$z_threshold),
               linetype = "dashed", colour = "grey50") +
    geom_vline(xintercept = c(-params$z_threshold, params$z_threshold),
               linetype = "dashed", colour = "grey50") +
    scale_colour_manual(values = c(
      setNames("#C44E52", paste(sp_a, "specific")),
      setNames("#4C72B0", paste(sp_b, "specific")),
      "Shared high" = "#DD8452",
      "Shared low"  = "#55A868",
      "Neutral"     = "#BBBBBB"
    )) +
    labs(title  = sprintf("%s vs %s", sp_a, sp_b),
         x      = sprintf("z(dN/dS) — %s", sp_a),
         y      = sprintf("z(dN/dS) — %s", sp_b),
         colour = NULL) +
    theme_classic(base_size = 11) +
    theme(legend.position = "bottom")
})

wrap_plots(scatter_plots, ncol = min(2, length(scatter_plots)))
```

### 5.2  UpSet plot — MKT methods × species

Each gene pair is annotated with the MKT tests that detected positive selection
in each species, using the boolean `pos_*` columns from the candidates tables.

```{r upset_methods, eval=run_cross_species, fig.width=10, fig.height=6, fig.cap="UpSet plot: overlap of MKT positive selection signals across species and tests (universe = gene pairs analysable in both species for both standard impMKT and branch-specific impMKT)."}
# Build upset dataset: one row per analyzable ortholog pair, boolean columns
# for each (species × test) combination.

build_upset_data <- function(pair_df) {
  sp_a <- pair_df$sp_a[1]; sp_b <- pair_df$sp_b[1]

  # Analyzable in both std_imp AND bs_imp
  az_a <- intersect(analyzable[[sp_a]]$std_imp_mkt, analyzable[[sp_a]]$bs_imp_mkt)
  az_b <- intersect(analyzable[[sp_b]]$std_imp_mkt, analyzable[[sp_b]]$bs_imp_mkt)

  df <- pair_df %>%
    filter(gene_a %in% az_a, gene_b %in% az_b)

  # Candidates from each species
  cand_a <- species_data[[sp_a]]$cand_all
  cand_b <- species_data[[sp_b]]$cand_all

  short_a <- short_name(sp_a); short_b <- short_name(sp_b)

  df %>%
    mutate(
      !!paste0(short_a, "_std_impMKT") :=
        gene_a %in% cand_a$gene_id[cand_a$pos_standard_impMKT == "TRUE"],
      !!paste0(short_a, "_bs_impMKT") :=
        gene_a %in% cand_a$gene_id[cand_a$pos_branch_specific_impMKT == "TRUE"],
      !!paste0(short_b, "_std_impMKT") :=
        gene_b %in% cand_b$gene_id[cand_b$pos_standard_impMKT == "TRUE"],
      !!paste0(short_b, "_bs_impMKT") :=
        gene_b %in% cand_b$gene_id[cand_b$pos_branch_specific_impMKT == "TRUE"]
    ) %>%
    mutate(across(matches("_std_impMKT|_bs_impMKT"), ~ replace_na(.x, FALSE)))
}

for (pair_name in names(pair_dfs)) {
  df   <- pair_dfs[[pair_name]]
  sp_a <- df$sp_a[1]; sp_b <- df$sp_b[1]
  short_a <- short_name(sp_a); short_b <- short_name(sp_b)

  upset_df  <- build_upset_data(df)
  upset_cols <- grep("_std_impMKT|_bs_impMKT", colnames(upset_df), value = TRUE)

  cat(sprintf("\n#### UpSet: %s vs %s  (%d pairs)\n", sp_a, sp_b, nrow(upset_df)))

  p <- upset(
    upset_df %>% select(all_of(upset_cols)) %>% as.data.frame(),
    intersect   = upset_cols,
    name        = "Method × species",
    width_ratio = 0.2,
    min_size    = 1,
    set_sizes   = upset_set_size()
  ) + ggtitle(sprintf("Positive selection methods — %s vs %s", sp_a, sp_b))
  print(p)
}
```

### 5.3  UpSet — MKT methods + high dN/dS

Same plot with additional `high dN/dS` sets for each species.

```{r upset_methods_highdnds, eval=run_cross_species, fig.width=12, fig.height=6, fig.cap="UpSet including high dN/dS categories."}
for (pair_name in names(pair_dfs)) {
  df   <- pair_dfs[[pair_name]]
  sp_a <- df$sp_a[1]; sp_b <- df$sp_b[1]
  short_a <- short_name(sp_a); short_b <- short_name(sp_b)

  upset_df   <- build_upset_data(df) %>%
    mutate(
      !!paste0(short_a, "_high_dNdS") := high_a,
      !!paste0(short_b, "_high_dNdS") := high_b
    ) %>%
    mutate(across(matches("_high_dNdS"), ~ replace_na(.x, FALSE)))

  upset_cols <- grep("_std_impMKT|_bs_impMKT|_high_dNdS",
                     colnames(upset_df), value = TRUE)

  p <- upset(
    upset_df %>% select(all_of(upset_cols)) %>% as.data.frame(),
    intersect   = upset_cols,
    name        = "Method × species",
    width_ratio = 0.2,
    min_size    = 1,
    set_sizes   = upset_set_size()
  ) + ggtitle(sprintf("MKT methods + high dN/dS — %s vs %s", sp_a, sp_b))
  print(p)
}
```

### 5.4  Enrichment of BS candidates in lineage-specific patterns

Fisher's exact test (one-sided greater): are branch-specific impMKT candidates
enriched among genes with lineage-specific high dN/dS?

```{r lineage_enrichment, eval=run_cross_species, results='asis'}
lineage_enrich <- lapply(names(pair_dfs), function(pair_name) {
  df   <- pair_dfs[[pair_name]]
  sp_a <- df$sp_a[1]; sp_b <- df$sp_b[1]

  run_fisher_lineage <- function(cand_genes, lineage_mask, universe) {
    cand  <- intersect(cand_genes, universe)
    bg    <- setdiff(universe, cand)
    a <- sum(cand %in% lineage_mask); b <- length(cand) - a
    c <- sum(bg   %in% lineage_mask); d <- length(bg)   - c
    mat <- matrix(c(a, b, c, d), nrow = 2,
                  dimnames = list(c("BS candidate", "Background"),
                                  c("Lineage-specific", "Other")))
    list(mat = mat, fisher = fisher.test(mat, alternative = "greater"))
  }

  list(
    pair = pair_name,
    sp_a = sp_a, sp_b = sp_b,
    a_specific = run_fisher_lineage(
      species_data[[sp_a]]$cand_bs$gene_id,
      df$gene_a[df$a_specific],
      df$gene_a
    ),
    b_specific = run_fisher_lineage(
      species_data[[sp_b]]$cand_bs$gene_id,
      df$gene_b[df$b_specific],
      df$gene_b
    )
  )
})
names(lineage_enrich) <- names(pair_dfs)

for (pair_name in names(lineage_enrich)) {
  r <- lineage_enrich[[pair_name]]
  cat(sprintf("\n#### %s — lineage-specific enrichment\n\n", pair_name))

  cat(sprintf("**%s-specific high dN/dS** (BS candidates enriched?)\n", r$sp_a))
  print(kbl(r$a_specific$mat) %>%
          kable_styling(bootstrap_options = c("striped", "condensed")))
  cat(sprintf("Fisher OR = %.3f  p = %.4f\n\n",
              r$a_specific$fisher$estimate, r$a_specific$fisher$p.value))

  cat(sprintf("**%s-specific high dN/dS** (BS candidates enriched?)\n", r$sp_b))
  print(kbl(r$b_specific$mat) %>%
          kable_styling(bootstrap_options = c("striped", "condensed")))
  cat(sprintf("Fisher OR = %.3f  p = %.4f\n\n",
              r$b_specific$fisher$estimate, r$b_specific$fisher$p.value))
}

# Write pair-level outputs
for (pair_name in names(pair_dfs)) {
  write_tsv(
    pair_dfs[[pair_name]] %>% select(-sp_a, -sp_b),
    file.path(params$output_dir, paste0("ortholog_pairs_", pair_name, ".tsv")),
    na = "NA"
  )
}
```

---

## Session info

```{r session_info}
sessionInfo()
```
