---
title: "`r paste0(params$experiment_name, ' — Expression vs Selection')`"
author: ""
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: yes
    toc_depth: 3
    toc_float: yes
    code_folding: hide
    df_print: paged
params:
  # ── Experiment identity ────────────────────────────────────────────────────
  experiment_name: "Experiment"
  count_file: ""                 # absolute path to count TSV
  # ── DESeq2 design ─────────────────────────────────────────────────────────
  control_condition: ""          # reference level (must appear in sample_conditions)
  sample_conditions: !r c()     # character vector — one element per sample column
  contrasts: !r list()          # named list: treatment_cond → reference_cond
                                 # if empty, auto: every unique condition vs control
  # ── MKT inputs (from branch_specific_MKT/render_analysis.R) ───────────────
  species_output_dirs: !r list() # named list: label → output_dir
  # ── Candidate gene set ─────────────────────────────────────────────────────
  # "bs_impMKT" = branch-specific imputed MKT candidates only
  # "bs_mkt"    = branch-specific standard MKT candidates only
  # "bs_any"    = union of bs_impMKT and bs_mkt (recommended)
  # "any"       = union of all 4 MKT tests
  candidate_type: "bs_any"
  # ── DE thresholds ──────────────────────────────────────────────────────────
  lfc_threshold: 0.5            # |log2FC| threshold for up/down classification
  padj_threshold: 0.05          # BH-adjusted p-value threshold
  # ── Output ─────────────────────────────────────────────────────────────────
  output_dir: "results/expression"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache   = FALSE,
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width  = 9,
  fig.height = 5
)
```

```{r packages, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(DESeq2)
  library(patchwork)
  library(kableExtra)
})
```

```{r helpers, include=FALSE}
# ── MKT filter (branch-specific analyzability) ────────────────────────────────
mk_filter_bs <- function(df) {
  df %>% filter(
    !is.na(num_pol_NS) & num_pol_NS >= 1,
    !is.na(num_pol_S)  & num_pol_S  >= 1,
    !is.na(Dn_branch_specific) & Dn_branch_specific >= 1,
    !is.na(Ds_branch_specific) & Ds_branch_specific >= 1,
    (num_pol_NS + num_pol_S) >= 5
  )
}

# ── Load and classify genes per species ───────────────────────────────────────
load_species_genes <- function(output_dir, candidate_type) {
  f <- file.path(output_dir, "branch_specific_MKT_results.tsv")
  if (!file.exists(f))
    stop("File not found: ", f,
         "\n  → Run branch_specific_MKT/render_analysis.R first.")

  mkt       <- read_tsv(f, show_col_types = FALSE)
  analyzable <- mk_filter_bs(mkt)

  get_positive_flag <- function(df, type) {
    switch(type,
      bs_impMKT = df$pos_branch_specific_impMKT %in% TRUE,
      bs_mkt    = df$pos_branch_specific_MKT    %in% TRUE,
      bs_any    = (df$pos_branch_specific_impMKT %in% TRUE) |
                  (df$pos_branch_specific_MKT    %in% TRUE),
      any       = (df$pos_standard_MKT           %in% TRUE) |
                  (df$pos_standard_impMKT         %in% TRUE) |
                  (df$pos_branch_specific_MKT     %in% TRUE) |
                  (df$pos_branch_specific_impMKT  %in% TRUE),
      stop("Unknown candidate_type: ", type)
    )
  }

  is_pos <- get_positive_flag(analyzable, candidate_type)

  clean_ids <- function(ids) {
    str_remove(na.omit(unique(ids)), "\\.[0-9]+$")
  }

  list(
    positive    = tibble(id = clean_ids(analyzable$aestivum_ortholog[is_pos])),
    neutral     = tibble(id = clean_ids(analyzable$aestivum_ortholog[!is_pos])),
    n_analyzable = nrow(analyzable)
  )
}

# ── Fisher test: selected vs neutral for one direction ─────────────────────────
fisher_one_direction <- function(all_results, pos_ids, neu_ids,
                                  lfc_thresh, padj_thresh, direction) {
  is_up   <- direction == "up"
  dir_lbl <- if (is_up) "upregulated" else "downregulated"

  summarise_de <- function(ids) {
    all_results %>%
      filter(id %in% ids) %>%
      group_by(condition) %>%
      summarise(
        n_total = n(),
        n_dir   = sum(
          padj < padj_thresh & !is.na(padj) &
          (if (is_up) log2FoldChange > lfc_thresh
           else        log2FoldChange < -lfc_thresh),
          na.rm = TRUE
        ),
        .groups = "drop"
      )
  }

  s_pos <- summarise_de(pos_ids)
  s_neu <- summarise_de(neu_ids)

  purrr::map_dfr(s_pos$condition, function(cond) {
    r_pos <- s_pos %>% filter(condition == cond)
    r_neu <- s_neu %>% filter(condition == cond)
    if (nrow(r_neu) == 0) return(NULL)

    mat <- matrix(
      c(r_pos$n_dir,  r_pos$n_total - r_pos$n_dir,
        r_neu$n_dir,  r_neu$n_total - r_neu$n_dir),
      nrow = 2, byrow = TRUE,
      dimnames = list(
        Selection  = c("Selected", "Neutral"),
        Expression = c(dir_lbl, paste0("not_", dir_lbl))
      )
    )
    ft    <- fisher.test(mat, alternative = "two.sided")
    stars <- case_when(
      ft$p.value < 0.001 ~ "***",
      ft$p.value < 0.01  ~ "**",
      ft$p.value < 0.05  ~ "*",
      TRUE               ~ "ns"
    )
    tibble(
      condition    = cond,
      direction    = dir_lbl,
      n_pos        = r_pos$n_total,
      n_dir_pos    = r_pos$n_dir,
      n_neu        = r_neu$n_total,
      n_dir_neu    = r_neu$n_dir,
      prop_pos     = r_pos$n_dir / r_pos$n_total,
      prop_neu     = r_neu$n_dir / r_neu$n_total,
      odds_ratio   = unname(ft$estimate),
      p_value      = ft$p.value,
      stars        = stars
    )
  })
}

# ── Bar-plot for one direction ─────────────────────────────────────────────────
plot_de_bars <- function(ft_df, sp_label, lfc_thresh, padj_thresh) {
  if (nrow(ft_df) == 0)
    return(ggplot() + theme_void() + labs(title = paste(sp_label, "— no data")))

  dir_lbl <- unique(ft_df$direction)
  df_long  <- ft_df %>%
    select(condition, prop_pos, prop_neu) %>%
    pivot_longer(c(prop_pos, prop_neu), names_to = "category", values_to = "proportion")

  annot <- ft_df %>%
    filter(stars != "ns") %>%
    left_join(
      df_long %>%
        group_by(condition) %>%
        summarise(y_top = max(proportion, na.rm = TRUE) * 1.15, .groups = "drop"),
      by = "condition"
    )

  ggplot(df_long, aes(x = condition, y = proportion, fill = category)) +
    geom_col(position = position_dodge(0.9), width = 0.8) +
    geom_text(
      data        = annot,
      aes(x = condition, y = y_top, label = stars),
      inherit.aes = FALSE, size = 5
    ) +
    scale_fill_manual(
      values = c(prop_pos = "firebrick3", prop_neu = "steelblue3"),
      labels = c(prop_pos = "Positive selection", prop_neu = "Neutral")
    ) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.15))) +
    labs(
      title = sprintf(
        "%s — %s (|LFC| > %.1f, padj < %.2f)",
        sp_label, dir_lbl, lfc_thresh, padj_thresh
      ),
      y    = "Proportion of DE genes",
      x    = NULL,
      fill = "Gene category"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1, vjust = 1),
      legend.position = "bottom",
      plot.title      = element_text(size = 11, face = "bold")
    )
}
```


# Experiment: `r params$experiment_name`

**Candidate type:** `r params$candidate_type`
| **|LFC| threshold:** `r params$lfc_threshold`
| **padj threshold:** `r params$padj_threshold`
| **Control:** `r params$control_condition`

---

## 1. Load expression data

```{r load-counts}
if (!nzchar(params$count_file))
  stop("params$count_file is empty. Set it in render_expression.R.")
if (!file.exists(params$count_file))
  stop("Count file not found: ", params$count_file)

expr_raw         <- read_tsv(params$count_file, show_col_types = FALSE)
colnames(expr_raw)[1] <- "id"

cat(sprintf("Count file   : %s\n", basename(params$count_file)))
cat(sprintf("Genes (rows) : %d\n", nrow(expr_raw)))
cat(sprintf("Samples      : %d\n", ncol(expr_raw) - 1))
```

## 2. Build count matrix

```{r count-matrix}
count_matrix <- expr_raw %>%
  tibble::column_to_rownames("id") %>%
  as.matrix() %>%
  round()
```

## 3. Experimental design

```{r coldata}
if (length(params$sample_conditions) == 0)
  stop("params$sample_conditions is empty. Set it in render_expression.R.")
if (length(params$sample_conditions) != ncol(count_matrix))
  stop(sprintf(
    "sample_conditions length (%d) != number of count matrix columns (%d).",
    length(params$sample_conditions), ncol(count_matrix)
  ))

coldata <- data.frame(
  condition = factor(params$sample_conditions),
  row.names = colnames(count_matrix)
)

if (nzchar(params$control_condition))
  coldata$condition <- relevel(coldata$condition, ref = params$control_condition)

stopifnot(all(colnames(count_matrix) == rownames(coldata)))
coldata
```

## 4. DESeq2 differential expression

```{r deseq2, results='hide'}
dds <- DESeqDataSetFromMatrix(
  countData = count_matrix,
  colData   = coldata,
  design    = ~condition
)
dds <- DESeq(dds)
```

### Contrasts

```{r contrasts}
# Derive contrasts: use params$contrasts if provided, else all conditions vs control
if (length(params$contrasts) > 0) {
  .contrasts <- params$contrasts
} else {
  if (!nzchar(params$control_condition))
    stop("Set params$control_condition when params$contrasts is empty.")
  .other_conds <- setdiff(levels(coldata$condition), params$control_condition)
  .contrasts <- setNames(
    as.list(rep(params$control_condition, length(.other_conds))),
    .other_conds
  )
}

extract_results <- function(dds, treatment, reference) {
  tryCatch(
    as.data.frame(results(dds, contrast = c("condition", treatment, reference))) %>%
      tibble::rownames_to_column("id") %>%
      mutate(condition = treatment) %>%
      select(id, condition, baseMean, log2FoldChange, lfcSE, stat, pvalue, padj),
    error = function(e) {
      warning(sprintf("Contrast '%s vs %s' failed: %s", treatment, reference, e$message))
      NULL
    }
  )
}

all_results <- bind_rows(purrr::imap(.contrasts, ~ extract_results(dds, .y, .x)))

cat(sprintf("Conditions tested  : %d\n", length(.contrasts)))
cat(sprintf("Total gene-results : %d\n", nrow(all_results)))
cat(sprintf("Genes with padj    : %d\n", sum(!is.na(all_results$padj))))
```

---

## 5. Per-species enrichment analysis

Gene sets are derived from `branch_specific_MKT_results.tsv` in each species output directory.
T. *aestivum* orthologs (`aestivum_ortholog` column) are used to map focal-species genes to
expression data.

```{r per-species, results='asis', fig.show='asis', fig.width=9, fig.height=5}
dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)

all_summaries    <- list()
species_genes_cache <- list()

for (.sp in names(params$species_output_dirs)) {
  cat(sprintf("\n\n### %s\n\n", .sp))

  # Load gene sets
  .genes <- tryCatch(
    load_species_genes(params$species_output_dirs[[.sp]], params$candidate_type),
    error = function(e) { cat("**ERROR:**", e$message, "\n"); NULL }
  )
  if (is.null(.genes)) next
  species_genes_cache[[.sp]] <- .genes

  .n_pos <- nrow(.genes$positive)
  .n_neu <- nrow(.genes$neutral)
  .n_in_expr_pos <- sum(.genes$positive$id %in% all_results$id)
  .n_in_expr_neu <- sum(.genes$neutral$id  %in% all_results$id)

  cat(sprintf(
    "**Analyzable:** %d | **Positive candidates:** %d (%d matched to expression data) | **Neutral:** %d (%d matched)\n\n",
    .genes$n_analyzable, .n_pos, .n_in_expr_pos, .n_neu, .n_in_expr_neu
  ))

  if (.n_in_expr_pos == 0) {
    cat("*No positive candidates overlap expression data — skipping species.*\n")
    next
  }

  # Fisher tests
  .ft_up   <- fisher_one_direction(
    all_results, .genes$positive$id, .genes$neutral$id,
    params$lfc_threshold, params$padj_threshold, "up"
  )
  .ft_down <- fisher_one_direction(
    all_results, .genes$positive$id, .genes$neutral$id,
    params$lfc_threshold, params$padj_threshold, "down"
  )

  # Combined result table
  .ft_all <- bind_rows(.ft_up, .ft_down)
  if (nrow(.ft_all) > 0) {
    print(
      kable(.ft_all,
            digits = 4, caption = sprintf("%s — Fisher test results", .sp)) %>%
        kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
    )
    cat("\n\n")
  }

  # Plots
  if (nrow(.ft_up) > 0) {
    print(plot_de_bars(.ft_up,   .sp, params$lfc_threshold, params$padj_threshold))
    cat("\n\n")
  }
  if (nrow(.ft_down) > 0) {
    print(plot_de_bars(.ft_down, .sp, params$lfc_threshold, params$padj_threshold))
    cat("\n\n")
  }

  # Save per-species results
  .out_file <- file.path(
    params$output_dir,
    sprintf("%s_%s_fisher_tests.tsv",
            params$experiment_name, gsub("[^A-Za-z0-9]", "_", .sp))
  )
  write_tsv(.ft_all, .out_file)

  # Accumulate for cross-species summary
  all_summaries[[.sp]] <- .ft_all %>%
    mutate(species = .sp, n_positive_candidates = .n_pos, n_neutral = .n_neu)
}
```

---

## 6. Cross-species summary

```{r cross-species-summary, results='asis'}
if (length(all_summaries) > 1) {
  .summary_tbl <- bind_rows(all_summaries) %>%
    select(species, condition, direction, n_pos, n_dir_pos, prop_pos,
           n_neu, n_dir_neu, prop_neu, odds_ratio, p_value, stars) %>%
    arrange(condition, direction, species)

  print(
    kable(.summary_tbl,
          digits = 4,
          caption = "Cross-species Fisher test summary") %>%
      kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
      column_spec(ncol(.summary_tbl), bold = TRUE)
  )

  write_tsv(.summary_tbl,
            file.path(params$output_dir,
                      paste0(params$experiment_name, "_cross_species_summary.tsv")))
} else {
  cat("*(only one species — cross-species table not generated)*\n")
}
```

---

## 7. Significant DE genes in selected candidates

```{r sig-genes-table}
if (length(species_genes_cache) > 0) {
  .sp1    <- names(species_genes_cache)[1]
  .genes1 <- species_genes_cache[[.sp1]]

  .sig_selected <- all_results %>%
    filter(
      id %in% .genes1$positive$id,
      !is.na(padj), padj < params$padj_threshold
    ) %>%
    arrange(padj) %>%
    head(50)

  if (nrow(.sig_selected) > 0) {
    cat(sprintf("Top significant DE genes in %s positive candidates:\n\n", .sp1))
    print(
      kable(.sig_selected, digits = 4) %>%
        kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
    )
  } else {
    cat("*No significant DE genes found among positive candidates.*\n")
  }
}
```

---

## Session info

```{r session-info}
cat(sprintf("Experiment   : %s\n", params$experiment_name))
cat(sprintf("Count file   : %s\n", basename(params$count_file)))
cat(sprintf("Candidate    : %s\n", params$candidate_type))
cat(sprintf("Control      : %s\n", params$control_condition))
cat(sprintf("Species      : %s\n", paste(names(params$species_output_dirs), collapse = ", ")))
cat(sprintf("Output dir   : %s\n", params$output_dir))
cat(sprintf("Date         : %s\n", Sys.Date()))
sessionInfo()
```
