---
title: "Coverage Analysis Report"
date: "`r Sys.Date()`"
params:
  output_dir: "."   # Path to the pipeline output directory produced by run_coverage_pipeline.sh
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
  pdf_document:
    toc: yes
    toc_depth: 2
---

<!--
  HOW TO USE
  ----------
  Option 1 — from the master pipeline wrapper (recommended):
    bash run_coverage_pipeline.sh --steps 10 --output /path/to/results

  Option 2 — render manually in R:
    rmarkdown::render(
      "Chromosome_coverage.Rmd",
      params = list(output_dir = "/path/to/results")
    )

  Option 3 — render from the command line:
    Rscript -e "rmarkdown::render('Chromosome_coverage.Rmd', \
      params = list(output_dir = '/path/to/results'))"

  The output_dir must be the same --output directory passed to
  run_coverage_pipeline.sh. It must contain at minimum the subdirectory
  step2_covered/ (generated by step 2). Other sections are optional and
  are skipped gracefully if their source files are absent.

  REQUIRED R PACKAGES (installed automatically below):
    ggplot2, dplyr, tidyr, scales, patchwork, knitr, kableExtra
-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 9,
  fig.height = 5
)

# ── Package management ────────────────────────────────────────────────────────
pkgs <- c("ggplot2", "dplyr", "tidyr", "scales", "patchwork", "knitr", "kableExtra")
new  <- pkgs[!pkgs %in% rownames(installed.packages())]
if (length(new)) {
  install.packages(new, repos = "https://cloud.r-project.org", quiet = TRUE)
}

suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(tidyr)
  library(scales)
  library(patchwork)
  library(knitr)
  library(kableExtra)
})

# ── Shared plot theme ─────────────────────────────────────────────────────────
theme_cov <- function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title       = element_text(face = "bold", size = 13),
      plot.subtitle    = element_text(colour = "grey45", size = 11),
      strip.text       = element_text(face = "bold"),
      axis.text.x      = element_text(angle = 45, hjust = 1)
    )
}

# ── Safe file reader: returns NULL if file does not exist ─────────────────────
read_pipeline_file <- function(path, sep = "\t", header = TRUE, ...) {
  if (!file.exists(path)) {
    message("File not found (step may not have run): ", basename(path))
    return(NULL)
  }
  read.table(path, sep = sep, header = header,
             stringsAsFactors = FALSE, check.names = FALSE, ...)
}

out_dir <- params$output_dir
```

```{r load_data, include=FALSE}
# ── Step 2 ────────────────────────────────────────────────────────────────────
f_summary    <- file.path(out_dir, "step2_covered/number_covered_position.txt")
f_contig_cov <- file.path(out_dir, "step2_covered/contig_percentage_covered_positions.txt")

summary_row <- read_pipeline_file(f_summary)
contig_cov  <- read_pipeline_file(f_contig_cov)

# ── Step 3 ────────────────────────────────────────────────────────────────────
f_bins   <- file.path(out_dir, "step3_coverage_bins/contig_coverage_summary.txt")
cov_bins <- read_pipeline_file(f_bins)

# ── Step 5 ────────────────────────────────────────────────────────────────────
f_depth     <- file.path(out_dir, "step5_mean_depth_per_contig/mean_depth_per_contig.txt")
mean_depth  <- read_pipeline_file(f_depth)

# ── Step 7 ────────────────────────────────────────────────────────────────────
f_heatmap <- file.path(out_dir,
  "step7_summary_tables/coverage_depth_summary_percentage.txt")
f_minpos  <- file.path(out_dir,
  "step7_summary_tables/coverage_depth_min_positions_counts.txt")

tbl_heatmap <- read_pipeline_file(f_heatmap)
tbl_minpos  <- read_pipeline_file(f_minpos)

# ── Step 8 (optional) ────────────────────────────────────────────────────────
f_sample   <- file.path(out_dir, "step8_whole_coverage/coverage_results.tsv")
sample_cov <- read_pipeline_file(f_sample)
```

---

> **Output directory:** `r out_dir`

---

# 1. Global Coverage Summary

```{r summary_table}
if (!is.null(summary_row)) {
  n_contigs     <- if (!is.null(contig_cov)) nrow(contig_cov) else NA
  n_samples_msg <- if (!is.null(contig_cov)) {
    # The depth matrix header gives the number of samples, but we can infer
    # from sample_cov if available
    if (!is.null(sample_cov)) paste(nrow(sample_cov), "samples") else "see step 1 log"
  } else "N/A"

  kable(
    data.frame(
      Metric = c(
        "Total genomic positions",
        "Covered positions",
        "Genome-wide coverage (%)",
        "Total contigs analysed",
        "Samples"
      ),
      Value = c(
        format(summary_row$total_positions,   big.mark = ",", scientific = FALSE),
        format(summary_row$covered_positions, big.mark = ",", scientific = FALSE),
        sprintf("%.2f %%", summary_row$percentage_covered),
        format(n_contigs, big.mark = ",", scientific = FALSE),
        n_samples_msg
      )
    ),
    col.names = c("Metric", "Value"),
    caption = paste("Filtering parameters used in step 2 are stored in:",
                    file.path(out_dir, "pipeline.log"))
  ) |>
    kable_styling(
      bootstrap_options = c("striped", "hover", "condensed"),
      full_width = FALSE,
      position = "left"
    )
} else {
  cat("Step 2 output not found. Run `bash run_coverage_pipeline.sh --steps 2 ...` first.\n")
}
```

---

# 2. Coverage Distribution across Contigs

## 2a. Coverage bins

Each bar shows how many contigs have that percentage of their positions covered
(at the minimum depth and minimum individuals set in step 2).

```{r coverage_bins_bar}
if (!is.null(cov_bins)) {
  cov_bins$pct <- as.numeric(sub("%", "", cov_bins$Threshold_coverage))

  ggplot(cov_bins,
         aes(x = factor(pct, levels = sort(unique(pct))),
             y = Count_of_contigs,
             fill = pct)) +
    geom_col(width = 0.82, colour = "white") +
    geom_text(aes(label = comma(Count_of_contigs)),
              vjust = -0.4, size = 3.5, colour = "grey30") +
    scale_fill_gradient(low = "#d73027", high = "#1a9850", guide = "none") +
    scale_y_continuous(labels = comma,
                       expand = expansion(mult = c(0, 0.13))) +
    labs(
      title    = "Number of contigs per coverage bin",
      subtitle = "A contig falls in a bin if that percentage of its positions are covered",
      x        = "Coverage bin (% of positions covered)",
      y        = "Number of contigs"
    ) +
    theme_cov()
} else {
  cat("Step 3 output not found. Run `--steps 3`.\n")
}
```

## 2b. Full distribution and cumulative curve

```{r contig_coverage_distribution}
if (!is.null(contig_cov)) {
  p_hist <- ggplot(contig_cov, aes(x = Percentage_covered)) +
    geom_histogram(binwidth = 5, fill = "#4575b4", colour = "white") +
    scale_x_continuous(limits = c(0, 100),
                       labels = function(x) paste0(x, "%")) +
    scale_y_continuous(labels = comma) +
    labs(
      title = "Histogram of per-contig coverage",
      x     = "% positions covered",
      y     = "Number of contigs"
    ) +
    theme_cov()

  p_cdf <- ggplot(contig_cov, aes(x = Percentage_covered)) +
    stat_ecdf(colour = "#d73027", linewidth = 1.1) +
    scale_x_continuous(limits = c(0, 100),
                       labels = function(x) paste0(x, "%")) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    labs(
      title    = "Cumulative distribution (CDF)",
      subtitle = "% of contigs with at least X% positions covered",
      x        = "% positions covered",
      y        = "Proportion of contigs"
    ) +
    theme_cov()

  p_hist + p_cdf
} else {
  cat("Step 2 output not found. Run `--steps 2`.\n")
}
```

**Interpretation:** A contig bimodality (peak near 0% and peak near 100%) is common in
RNA-seq transcriptomes — some transcripts are simply not expressed in the sampled
conditions. Contigs in the middle range may reflect partial coverage due to expression
level, mapping quality, or reference incompleteness.

---

# 3. Mean Sequencing Depth per Contig

```{r mean_depth_distribution}
if (!is.null(mean_depth)) {
  med_d  <- median(mean_depth$mean_depth, na.rm = TRUE)
  mean_d <- mean(mean_depth$mean_depth,   na.rm = TRUE)
  q25    <- quantile(mean_depth$mean_depth, 0.25, na.rm = TRUE)
  q75    <- quantile(mean_depth$mean_depth, 0.75, na.rm = TRUE)

  p_main <- ggplot(mean_depth, aes(x = mean_depth)) +
    geom_histogram(bins = 60, fill = "#74add1", colour = "white") +
    geom_vline(xintercept = med_d, linetype = "dashed",
               colour = "#d73027", linewidth = 0.9) +
    geom_vline(xintercept = mean_d, linetype = "dotted",
               colour = "#1a9850", linewidth = 0.9) +
    annotate("text", x = med_d, y = Inf, vjust = 1.8, hjust = -0.12,
             label = paste0("Median\n", round(med_d, 1), "×"),
             colour = "#d73027", size = 3.5) +
    annotate("text", x = mean_d, y = Inf, vjust = 1.8,
             hjust = if (mean_d > med_d) -0.12 else 1.1,
             label = paste0("Mean\n", round(mean_d, 1), "×"),
             colour = "#1a9850", size = 3.5) +
    scale_x_continuous(labels = comma) +
    scale_y_continuous(labels = comma) +
    labs(
      title    = "Distribution of mean depth per contig",
      subtitle = "Computed from covered positions only (steps 4–5)",
      x        = "Mean depth (reads per covered position)",
      y        = "Number of contigs"
    ) +
    theme_cov()

  p_box <- ggplot(mean_depth, aes(x = "", y = mean_depth)) +
    geom_violin(fill = "#74add1", colour = "#4575b4", alpha = 0.7) +
    geom_boxplot(width = 0.15, fill = "white", colour = "grey30",
                 outlier.size = 0.5) +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "Mean depth (×)") +
    theme_cov() +
    theme(axis.text.x = element_blank())

  p_main + p_box + plot_layout(widths = c(3, 1))
} else {
  cat("Step 5 output not found. Run `--steps 5`.\n")
}
```

```{r mean_depth_table, echo=FALSE}
if (!is.null(mean_depth)) {
  kable(
    data.frame(
      Statistic = c("Min", "Q25", "Median", "Mean", "Q75", "Max"),
      Value = c(
        round(min(mean_depth$mean_depth,    na.rm = TRUE), 1),
        round(q25, 1),
        round(med_d, 1),
        round(mean_d, 1),
        round(q75, 1),
        round(max(mean_depth$mean_depth,    na.rm = TRUE), 1)
      )
    ),
    caption = "Summary statistics of mean depth per contig (×)"
  ) |>
    kable_styling(bootstrap_options = c("striped", "hover"),
                  full_width = FALSE, position = "left")
}
```

---

# 4. Coverage × Depth Threshold Matrix

What percentage of contigs pass both a minimum % coverage **and** a minimum depth
threshold simultaneously? Use this heatmap to choose appropriate filtering parameters
for your analysis.

```{r coverage_depth_heatmap}
if (!is.null(tbl_heatmap)) {
  # Rename the first column (contains special characters in the original file)
  colnames(tbl_heatmap)[1] <- "Cov_threshold"

  # Parse "5%" → 5
  tbl_heatmap$cov_num <- as.numeric(sub("%", "", tbl_heatmap$Cov_threshold))

  depth_cols <- setdiff(colnames(tbl_heatmap), c("Cov_threshold", "cov_num"))

  long <- tbl_heatmap |>
    select(-Cov_threshold) |>
    pivot_longer(-cov_num,
                 names_to  = "Depth_threshold",
                 values_to = "Pct_contigs")

  long$depth_num       <- as.numeric(sub("X", "", long$Depth_threshold))
  long$Depth_threshold <- factor(long$Depth_threshold,
                                  levels = paste0(sort(unique(long$depth_num)), "X"))
  long$Pct_contigs     <- as.numeric(long$Pct_contigs)

  ggplot(long, aes(x = Depth_threshold,
                   y = factor(cov_num, levels = sort(unique(cov_num))),
                   fill = Pct_contigs)) +
    geom_tile(colour = "white", linewidth = 0.5) +
    geom_text(aes(label = sprintf("%.1f", Pct_contigs)), size = 3) +
    scale_fill_gradient2(
      low      = "#d73027",
      mid      = "#ffffbf",
      high     = "#1a9850",
      midpoint = 50,
      limits   = c(0, 100),
      name     = "% of\ncontigs"
    ) +
    scale_y_discrete(labels = function(x) paste0(x, "%")) +
    labs(
      title    = "% of contigs meeting coverage AND depth thresholds",
      subtitle = "Row = minimum % positions covered per contig  |  Column = minimum depth per covered position",
      x        = "Minimum depth threshold",
      y        = "Minimum % positions covered"
    ) +
    theme_cov() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
} else {
  cat("Step 7 output not found. Run `--steps 7`.\n")
}
```

**How to read this:** The top-right cell (high coverage %, high depth) shows what proportion
of contigs survive the strictest filters. Green cells (high %) indicate a well-covered dataset.

---

# 5. Number of Contigs with Minimum Positions at Each Depth

How many contigs retain a minimum number of positions when a depth cut-off is applied?
Each line corresponds to a different minimum number of positions required.

```{r min_positions_lines}
if (!is.null(tbl_minpos)) {
  colnames(tbl_minpos)[1] <- "Min_positions"

  long2 <- tbl_minpos |>
    pivot_longer(-Min_positions,
                 names_to  = "Depth_threshold",
                 values_to = "N_contigs")

  long2$depth_num       <- as.numeric(sub("X", "", long2$Depth_threshold))
  long2$Depth_threshold <- factor(long2$Depth_threshold,
                                   levels = paste0(sort(unique(long2$depth_num)), "X"))
  long2$N_contigs       <- as.numeric(long2$N_contigs)
  long2$Min_label       <- factor(
    paste0("≥", long2$Min_positions, " pos"),
    levels = paste0("≥", sort(unique(long2$Min_positions)), " pos")
  )

  ggplot(long2, aes(x = Depth_threshold, y = N_contigs,
                    colour = Min_label, group = Min_label)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2.8) +
    scale_y_continuous(labels = comma) +
    scale_colour_brewer(palette = "Set1",
                        name = "Min. covered\npositions") +
    labs(
      title    = "Contigs passing a minimum-position filter at each depth",
      subtitle = "Each line = a different minimum number of positions required at that depth",
      x        = "Minimum depth threshold",
      y        = "Number of contigs retained"
    ) +
    theme_cov() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
} else {
  cat("Step 7 output not found. Run `--steps 7`.\n")
}
```

---

# 6. Per-Sample Sequencing Coverage

> This section is only shown when step 8 (whole-genome coverage estimation) has been run.

```{r sample_coverage_check, results='asis'}
if (is.null(sample_cov)) {
  cat("> **Step 8 output not found.** To generate this section, re-run the pipeline with\n",
      "> `--steps 8 --bam-dir /path/bams --ref /path/ref.fa`.\n")
}
```

```{r sample_coverage_plots}
if (!is.null(sample_cov)) {
  sample_cov <- sample_cov[order(sample_cov$Coverage), ]
  sample_cov$Sample <- factor(sample_cov$Sample, levels = sample_cov$Sample)

  p_cov <- ggplot(sample_cov,
                  aes(x = Coverage, y = Sample, fill = Coverage)) +
    geom_col() +
    geom_text(aes(label = sprintf("%.1f×", Coverage)),
              hjust = -0.1, size = 3.5) +
    scale_fill_gradient(low = "#fdae61", high = "#1a9850", guide = "none") +
    scale_x_continuous(expand = expansion(mult = c(0, 0.18))) +
    labs(title = "Coverage per sample", x = "Coverage (×)", y = NULL) +
    theme_cov() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

  p_reads <- ggplot(sample_cov,
                    aes(x = Reads_mapped / 1e6, y = Sample, fill = Coverage)) +
    geom_col() +
    scale_fill_gradient(low = "#fdae61", high = "#1a9850", guide = "none") +
    scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
    labs(title = "Reads mapped per sample",
         x = "Reads mapped (millions)", y = NULL) +
    theme_cov() +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))

  p_cov + p_reads &
    theme(axis.text.y = element_text(size = 9))
}
```

```{r sample_coverage_table}
if (!is.null(sample_cov)) {
  df <- sample_cov[order(-sample_cov$Coverage), ] |>
    mutate(
      Reads_mapped     = format(Reads_mapped, big.mark = ",", scientific = FALSE),
      Mean_Read_Length = paste0(Mean_Read_Length, " bp"),
      Coverage         = sprintf("%.2f×", Coverage)
    )

  kable(df,
        col.names = c("Sample", "Reads mapped", "Mean read length", "Coverage"),
        caption   = "Samples ordered by coverage (highest first)") |>
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE)
}
```

---

*Report generated by `Chromosome_coverage.Rmd` — Coverage Analysis Pipeline.*
*Rendered: `r format(Sys.time(), "%Y-%m-%d %H:%M:%S")`*
*Output directory: `r out_dir`*
