---
title: "Recombination Rate and Positive Selection"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    fig_caption: true
    self_contained: true
params:
  # Named list: species_label → config list.
  # Each entry must have:
  #   output_dir      : output_dir used in render_analysis.R (for MKT results)
  #   gene_coords_bed : 4-column BED with genomic coordinates of all genes
  #                     Columns (no header): chr  start  end  gene_id
  #   rec_map_bed     : 4-column BED with recombination rate intervals
  #                     Columns (no header): chr  start  end  rec_rate
  species_configs: !r list()
  # Candidate set used as the binary response variable ("selected"):
  #   "bs"      = branch-specific MKT or branch-specific impMKT (recommended)
  #   "any"     = positive in any of the 4 MKT tests
  candidate_type: "bs"
  # MKT filter applied to define the statistical universe (analyzable genes):
  # Pn + Ps >= min_poly AND all four cells (Pn, Ps, Dn, Ds) >= min_count
  min_poly:  5
  min_count: 1
  # z-score threshold for sensitivity analysis (exclude |distance_z| > threshold)
  distance_z_threshold: 3
  # Scaling factor applied to rec_rate before modelling (default 1e6 = per Mb)
  rec_scale: 1000000
  # Output directory for TSV result tables
  output_dir: "results/recombination"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", fig.width = 10, fig.height = 6)

# Compute a human-readable rec_rate axis label once, used in all plots.
# With rec_scale = 1e6 (default) and Glemin et al. data in cM/bp,
# rec_rate_scaled is in cM/Mb.
rec_unit_label <- if (params$rec_scale == 1e6) "cM/Mb" else
                  sprintf("rec_rate × %g", params$rec_scale)
```

```{r libraries}
library(tidyverse)
library(GenomicRanges)
library(IRanges)
library(S4Vectors)
library(kableExtra)
library(patchwork)
```

This analysis tests whether genes under branch-specific positive selection are
enriched in genomic regions with high recombination rates. The question is
motivated by the expected positive correlation between recombination and
adaptive evolution (Hill-Robertson interference, reduced background selection).

**Statistical model:** binomial logistic GLM with binary selection status as the
response. Assignment uncertainty (distance from gene to the nearest
recombination interval) enters the model as a standardised covariate
(`distance_z`) rather than being handled by separate sub-analyses.

$$\text{logit}(P[\text{selected}]) = \beta_0 + \beta_1 \cdot \text{rec\_rate\_scaled} + \beta_2 \cdot \text{distance\_z}$$

---

## 1  Parameter validation and input loading

```{r validate_params}
if (length(params$species_configs) == 0)
  stop("params$species_configs is empty. Edit render_recombinaison.R and provide ",
       "at least one species.")

sp_labels  <- names(params$species_configs)
n_species  <- length(sp_labels)
cand_type  <- params$candidate_type

cat(sprintf("Species         : %s\n", paste(sp_labels, collapse = ", ")))
cat(sprintf("Candidate type  : %s\n", cand_type))
cat(sprintf("MKT filter      : Pn+Ps >= %d, all cells >= %d\n",
            params$min_poly, params$min_count))
cat(sprintf("distance_z cutoff (sensitivity): %.1f\n", params$distance_z_threshold))
cat(sprintf("rec_scale       : %g\n", params$rec_scale))
cat(sprintf("Output dir      : %s\n", params$output_dir))

dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)
```

### 1.1  MKT filter helper

Genes are **analyzable** when they pass the McDonald-Kreitman filter for the
branch-specific tests. Only these genes enter the recombination analysis.

```{r mk_filter_helper}
mk_filter_bs <- function(stats_df, min_poly, min_count) {
  with(stats_df,
    !is.na(num_pol_NS)         & num_pol_NS         >= min_count &
    !is.na(num_pol_S)          & num_pol_S          >= min_count &
    !is.na(Dn_branch_specific) & Dn_branch_specific >= min_count &
    !is.na(Ds_branch_specific) & Ds_branch_specific >= min_count &
    (num_pol_NS + num_pol_S)   >= min_poly
  )
}
```

### 1.2  Load and assemble per-species data

For each species the following inputs are read:

| Source | Description |
|--------|-------------|
| `branch_specific_MKT_results.tsv` | MKT statistics — defines analyzable universe |
| `candidates_branch_specific.tsv` or `candidates_all.tsv` | Positive candidates → binary response |
| `gene_coords_bed` | Gene genomic coordinates (chr, start, end, gene_id) |
| `rec_map_bed` | Recombination rate intervals (chr, start, end, rec_rate) |

The `selection` column (0/1) is derived from whether a gene appears in the
candidate table. The analyzable filter restricts the universe to genes with
valid branch-specific divergence counts.

```{r load_species}
load_species <- function(sp, cfg, cand_type, min_poly, min_count) {

  # ── MKT results ──────────────────────────────────────────────────────────────
  stats_file <- file.path(cfg$output_dir, "branch_specific_MKT_results.tsv")
  if (cand_type == "bs") {
    cand_file <- file.path(cfg$output_dir, "candidates_branch_specific.tsv")
  } else {
    cand_file <- file.path(cfg$output_dir, "candidates_all.tsv")
  }

  for (f in c(stats_file, cand_file, cfg$gene_coords_bed, cfg$rec_map_bed)) {
    if (!file.exists(f))
      stop(sprintf("[%s] File not found: %s\nRun render_analysis.R first,\n",
                   sp, f),
           "then check that gene_coords_bed and rec_map_bed paths are correct.")
  }

  stats <- read_tsv(stats_file,
                    col_types = cols(.default = col_double(),
                                     gene_id  = col_character(),
                                     ortholog = col_character()),
                    na = c("NA", "NaN", ""))

  cands <- read_tsv(cand_file,
                    col_types = cols(.default = col_character()),
                    na = c("NA", "NaN", ""))

  # Analyzable genes = genes passing branch-specific MKT filter
  analyzable_ids <- stats$gene_id[mk_filter_bs(stats, min_poly, min_count)]

  # Positive candidate gene IDs
  candidate_ids <- unique(cands$gene_id)

  # ── Gene coordinates BED ─────────────────────────────────────────────────────
  genes_raw <- read_tsv(cfg$gene_coords_bed,
                        col_names = c("chr", "start", "end", "gene_id"),
                        col_types = cols(chr   = col_character(),
                                         start = col_integer(),
                                         end   = col_integer(),
                                         gene_id = col_character()),
                        na = c("NA", ""))

  # Restrict to analyzable genes and add binary selection status
  genes <- genes_raw %>%
    filter(gene_id %in% analyzable_ids) %>%
    mutate(selection = as.integer(gene_id %in% candidate_ids))

  # ── Recombination map BED ────────────────────────────────────────────────────
  rec <- read_tsv(cfg$rec_map_bed,
                  col_names = c("chr", "start", "end", "rec_rate"),
                  col_types = cols(chr      = col_character(),
                                   start    = col_integer(),
                                   end      = col_integer(),
                                   rec_rate = col_double()),
                  na = c("NA", ""))
  rec <- rec[!is.na(rec$rec_rate), , drop = FALSE]

  list(
    sp            = sp,
    cfg           = cfg,
    stats         = stats,
    analyzable_ids = analyzable_ids,
    candidate_ids  = candidate_ids,
    genes         = genes,
    rec           = rec
  )
}

raw_data <- setNames(
  lapply(sp_labels, function(sp)
    load_species(sp, params$species_configs[[sp]],
                 cand_type, params$min_poly, params$min_count)),
  sp_labels
)
```

```{r input_summary}
summary_rows <- lapply(sp_labels, function(sp) {
  d <- raw_data[[sp]]
  tibble(
    Species           = sp,
    Genes_total       = nrow(raw_data[[sp]]$stats),
    Analyzable        = length(d$analyzable_ids),
    Candidates        = length(d$candidate_ids),
    Genes_in_BED      = nrow(raw_data[[sp]]$genes),
    Selected_in_BED   = sum(raw_data[[sp]]$genes$selection),
    Rec_intervals     = nrow(d$rec),
    Chrs_genes        = n_distinct(d$genes$chr),
    Chrs_rec          = n_distinct(d$rec$chr)
  )
})

bind_rows(summary_rows) %>%
  kbl(caption = "Input data summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 2  GenomicRanges: assign recombination rate to each gene

Each gene receives a recombination rate by:
1. **Overlap**: the mean rec rate of all overlapping intervals (preferred)
2. **Nearest**: the rec rate of the nearest non-overlapping interval
3. **Population mean**: mean across all intervals (fallback when no nearest found)

The distance from each gene to its assigned interval is computed and log-
transformed then standardised (`distance_z`) to enter the model as a continuous
covariate. For overlapping genes `distance_z = 0`.

Only genes on chromosomes that have recombination data are retained.

```{r assign_recombination}
assign_rec <- function(sp, data_list, rec_scale) {
  d     <- data_list[[sp]]
  genes <- d$genes
  rec   <- d$rec

  chrs_rec   <- unique(rec$chr)
  chrs_genes <- unique(genes$chr)
  chrs_no_rec <- setdiff(chrs_genes, chrs_rec)

  if (length(chrs_no_rec) > 0)
    warning(sprintf("[%s] %d chromosomes in gene BED have NO recombination data: %s",
                    sp, length(chrs_no_rec),
                    paste(chrs_no_rec, collapse = ", ")))

  genes <- genes[genes$chr %in% chrs_rec, , drop = FALSE]

  if (nrow(genes) == 0)
    stop(sprintf("[%s] No genes remain after restricting to chromosomes with ",
                 sp),
         "recombination data. Check chromosome naming between BEDs.")

  gr_genes <- GRanges(seqnames = genes$chr,
                      ranges   = IRanges(start = genes$start, end = genes$end),
                      gene_id   = genes$gene_id,
                      selection = genes$selection,
                      gene_idx  = seq_len(nrow(genes)))

  gr_rec   <- GRanges(seqnames = rec$chr,
                      ranges   = IRanges(start = rec$start, end = rec$end),
                      rec_rate = rec$rec_rate)

  # ── Step 1: overlap assignment ───────────────────────────────────────────────
  overlaps <- findOverlaps(gr_genes, gr_rec, ignore.strand = TRUE)
  if (length(overlaps) > 0) {
    ovl_df <- data.frame(
      gene_idx = queryHits(overlaps),
      rec_rate = mcols(gr_rec)$rec_rate[subjectHits(overlaps)],
      stringsAsFactors = FALSE
    )
    ovl_df <- aggregate(rec_rate ~ gene_idx, data = ovl_df, FUN = mean)
    n_per_gene <- as.data.frame(table(queryHits(overlaps)), stringsAsFactors = FALSE)
    names(n_per_gene) <- c("gene_idx", "n_overlaps")
    ovl_df <- merge(ovl_df, n_per_gene, by = "gene_idx", all.x = TRUE)
    ovl_df$assignment_method        <- "overlap"
    ovl_df$distance_to_interval_kb  <- 0
    genes_with_overlap <- unique(ovl_df$gene_idx)
  } else {
    ovl_df <- data.frame()
    genes_with_overlap <- integer(0)
  }

  # ── Step 2: nearest assignment ───────────────────────────────────────────────
  genes_no_overlap <- setdiff(seq_along(gr_genes), genes_with_overlap)
  if (length(genes_no_overlap) > 0) {
    nearest_idx  <- nearest(gr_genes[genes_no_overlap], gr_rec, ignore.strand = TRUE)
    dist_obj     <- distanceToNearest(gr_genes[genes_no_overlap], gr_rec, ignore.strand = TRUE)
    has_nearest  <- !is.na(nearest_idx)

    if (any(has_nearest)) {
      near_df <- data.frame(
        gene_idx               = genes_no_overlap[has_nearest],
        rec_rate               = mcols(gr_rec)$rec_rate[nearest_idx[has_nearest]],
        distance_to_interval_kb = mcols(dist_obj)$distance[has_nearest] / 1000,
        assignment_method      = "nearest",
        stringsAsFactors       = FALSE
      )
    } else {
      near_df <- data.frame()
    }

    # Step 3: population-mean fallback
    genes_no_nearest <- genes_no_overlap[!has_nearest]
    if (length(genes_no_nearest) > 0) {
      fallback_df <- data.frame(
        gene_idx               = genes_no_nearest,
        rec_rate               = mean(rec$rec_rate, na.rm = TRUE),
        distance_to_interval_kb = NA_real_,
        assignment_method      = "population_mean",
        stringsAsFactors       = FALSE
      )
      near_df <- bind_rows(near_df, fallback_df)
    }
  } else {
    near_df <- data.frame()
  }

  # ── Assemble combined data frame ─────────────────────────────────────────────
  # Use select() rather than [, c(...)] so empty data frames (0 rows, 0 cols)
  # don't crash — bind_rows() handles empty tibbles cleanly.
  cols_keep <- c("gene_idx", "rec_rate", "assignment_method", "distance_to_interval_kb")
  safe_sel  <- function(df) {
    if (nrow(df) == 0)
      return(tibble(gene_idx = integer(), rec_rate = double(),
                    assignment_method = character(), distance_to_interval_kb = double()))
    df[, cols_keep]
  }

  combined <- bind_rows(safe_sel(ovl_df), safe_sel(near_df)) %>%
    arrange(gene_idx) %>%
    mutate(
      gene_id          = mcols(gr_genes)$gene_id[gene_idx],
      selection        = as.integer(mcols(gr_genes)$selection[gene_idx]),
      chr              = as.character(seqnames(gr_genes))[gene_idx],
      pos              = (start(gr_genes) + end(gr_genes))[gene_idx] / 2,
      rec_rate_scaled  = rec_rate * rec_scale,
      distance_z       = as.numeric(scale(log1p(distance_to_interval_kb)))
    )

  list(
    combined       = combined,
    gr_genes       = gr_genes,
    gr_rec         = gr_rec,
    chrs_no_rec    = chrs_no_rec
  )
}

assigned <- setNames(
  lapply(sp_labels, function(sp) assign_rec(sp, raw_data, params$rec_scale)),
  sp_labels
)
```

```{r assignment_summary}
assign_tbl <- bind_rows(lapply(sp_labels, function(sp) {
  df <- assigned[[sp]]$combined
  tibble(
    Species         = sp,
    n_analyzed      = nrow(df),
    n_selected      = sum(df$selection == 1, na.rm = TRUE),
    n_overlap       = sum(df$assignment_method == "overlap"),
    n_nearest       = sum(df$assignment_method == "nearest"),
    n_pop_mean      = sum(df$assignment_method == "population_mean"),
    pct_direct      = round(100 * sum(df$assignment_method == "overlap") / nrow(df), 1),
    median_rec      = round(median(df$rec_rate, na.rm = TRUE), 6),
    NAs_distance_z  = sum(is.na(df$distance_z))
  )
}))

assign_tbl %>%
  kbl(caption = "Recombination rate assignment summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

```{r plot_assignment_methods, fig.cap="Recombination rate distributions by assignment method. Genes with direct overlap assignments are in blue; nearest-interval and population-mean fallbacks in other colours."}
plots_assign <- lapply(sp_labels, function(sp) {
  assigned[[sp]]$combined %>%
    ggplot(aes(x = rec_rate_scaled, fill = assignment_method,
               colour = assignment_method)) +
    geom_histogram(bins = 40, alpha = 0.7, position = "identity") +
    scale_fill_manual(values = c(overlap = "#4C72B0", nearest = "#DD8452",
                                  population_mean = "#8C8C8C")) +
    scale_colour_manual(values = c(overlap = "#4C72B0", nearest = "#DD8452",
                                    population_mean = "#8C8C8C"), guide = "none") +
    labs(title = sp,
         x     = paste("Recombination rate (", rec_unit_label, ")"),
         y     = "Genes", fill  = "Assignment") +
    theme_classic(base_size = 11) +
    theme(legend.position = "bottom")
})

wrap_plots(plots_assign, ncol = min(2, n_species))
```

---

## 3  Recombination rate vs selection status

```{r plot_rec_vs_selection, fig.cap="Recombination rate distributions for selected vs non-selected genes."}
plots_rec <- lapply(sp_labels, function(sp) {
  df <- assigned[[sp]]$combined %>%
    mutate(group = if_else(selection == 1, "Candidate", "Background"))

  ggplot(df, aes(x = rec_rate_scaled, fill = group, colour = group)) +
    geom_density(alpha = 0.45, adjust = 1.5) +
    scale_fill_manual(values   = c(Candidate = "#C44E52", Background = "#4C72B0")) +
    scale_colour_manual(values = c(Candidate = "#C44E52", Background = "#4C72B0"),
                        guide  = "none") +
    labs(title = sp,
         x     = paste("Recombination rate (", rec_unit_label, ")"),
         y     = "Density", fill = NULL) +
    theme_classic(base_size = 11) +
    theme(legend.position = "bottom")
})

wrap_plots(plots_rec, ncol = min(2, n_species))
```

```{r rec_medians}
med_tbl <- bind_rows(lapply(sp_labels, function(sp) {
  assigned[[sp]]$combined %>%
    mutate(group = if_else(selection == 1, "Candidate", "Background")) %>%
    group_by(Species = sp, group) %>%
    summarise(n = n(),
              median_rec = signif(median(rec_rate_scaled, na.rm = TRUE), 4),
              mean_rec   = signif(mean(rec_rate_scaled,   na.rm = TRUE), 4),
              .groups = "drop")
}))

med_tbl %>%
  kbl(caption = "Median recombination rate — candidates vs background") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 4  Binomial logistic GLM

### 4.1  Main model

$$\text{logit}(P[\text{selected}]) = \beta_0 + \beta_1 \cdot \text{rec\_rate\_scaled} + \beta_2 \cdot \text{distance\_z}$$

- `rec_rate_scaled` tests whether recombination rate predicts positive selection
  probability after accounting for assignment uncertainty.
- `distance_z` captures the effect of assignment uncertainty; a large positive
  coefficient indicates that genes far from measured intervals behave differently.

```{r glm_main}
fit_models <- function(sp) {
  df <- assigned[[sp]]$combined %>%
    filter(!is.na(selection), !is.na(rec_rate_scaled))

  main        <- glm(selection ~ rec_rate_scaled + distance_z,
                     family = binomial(link = "logit"), data = df)
  interaction <- glm(selection ~ rec_rate_scaled * distance_z,
                     family = binomial(link = "logit"), data = df)
  list(main = main, interaction = interaction, data = df)
}

models <- setNames(lapply(sp_labels, fit_models), sp_labels)

for (sp in sp_labels) {
  cat(sprintf("\n### Main model — %s\n\n", sp))
  print(summary(models[[sp]]$main))
}
```

### 4.2  Interaction model and LRT

```{r glm_interaction}
for (sp in sp_labels) {
  cat(sprintf("\n### Interaction model — %s\n\n", sp))
  print(summary(models[[sp]]$interaction))
  cat(sprintf("\nLikelihood-ratio test (main vs interaction) — %s\n", sp))
  print(anova(models[[sp]]$main, models[[sp]]$interaction, test = "Chisq"))
}
```

### 4.3  Model coefficient table

```{r glm_coef_table}
coef_tbl <- bind_rows(lapply(sp_labels, function(sp) {
  coef_mat <- coef(summary(models[[sp]]$main))
  as.data.frame(coef_mat) %>%
    rownames_to_column("term") %>%
    mutate(
      Species    = sp,
      odds_ratio = round(exp(Estimate), 4),
      Estimate   = round(Estimate, 5),
      `Std. Error` = round(`Std. Error`, 5),
      `z value`    = round(`z value`, 3),
      `Pr(>|z|)`   = signif(`Pr(>|z|)`, 4)
    ) %>%
    select(Species, term, Estimate, `Std. Error`, `z value`, `Pr(>|z|)`, odds_ratio)
}))

coef_tbl %>%
  kbl(caption = "Main GLM coefficients (all species)") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(coef_tbl$`Pr(>|z|)` < 0.05), bold = TRUE)
```

```{r plot_coef, fig.cap="Estimated odds ratios (OR) for the recombination rate coefficient. OR > 1: positive relationship between recombination and candidate status."}
rec_coefs <- filter(coef_tbl, term == "rec_rate_scaled")

ggplot(rec_coefs, aes(x = Species, y = odds_ratio, ymin = exp(Estimate - 1.96 * `Std. Error`),
                       ymax = exp(Estimate + 1.96 * `Std. Error`))) +
  geom_pointrange(colour = "#4C72B0", size = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  labs(title = "Odds ratio for recombination rate (95% CI)",
       x = NULL, y = "Odds ratio (rec_rate_scaled)") +
  theme_classic(base_size = 11)
```

---

## 5  Sensitivity analysis — distance-z filter

Genes with `|distance_z|` > `r params$distance_z_threshold` are excluded.
This tests whether the recombination coefficient is driven by genes assigned
from very distant recombination intervals.

```{r sensitivity_analysis}
sens_results <- bind_rows(lapply(sp_labels, function(sp) {
  df_full <- models[[sp]]$data
  df_filt <- filter(df_full, !is.na(distance_z),
                    abs(distance_z) <= params$distance_z_threshold)

  n_total    <- nrow(df_full)
  n_retained <- nrow(df_filt)
  n_excluded <- n_total - n_retained

  cat(sprintf("\n[%s] distance_z filter (|z| <= %.1f):\n",
              sp, params$distance_z_threshold))
  cat(sprintf("  Total: %d  |  Retained: %d  |  Excluded: %d (%.1f%%)\n",
              n_total, n_retained, n_excluded, 100 * n_excluded / n_total))

  if (n_retained == 0) {
    warning(sprintf("[%s] No genes retained after sensitivity filter.", sp))
    return(NULL)
  }

  m_filt  <- glm(selection ~ rec_rate_scaled + distance_z,
                 family = binomial(link = "logit"), data = df_filt)
  cf_full <- coef(summary(models[[sp]]$main))["rec_rate_scaled", ]
  cf_filt <- coef(summary(m_filt))["rec_rate_scaled", ]

  tibble(
    species          = sp,
    n_total          = n_total,
    n_retained       = n_retained,
    n_excluded       = n_excluded,
    full_estimate    = round(unname(cf_full["Estimate"]),      5),
    full_OR          = round(exp(unname(cf_full["Estimate"])), 4),
    full_p           = signif(unname(cf_full["Pr(>|z|)"]),     4),
    filtered_estimate = round(unname(cf_filt["Estimate"]),      5),
    filtered_OR       = round(exp(unname(cf_filt["Estimate"])), 4),
    filtered_p        = signif(unname(cf_filt["Pr(>|z|)"]),     4)
  )
}))

sens_results %>%
  kbl(caption = sprintf("Full model vs sensitivity-filtered model (|distance_z| <= %.1f)",
                        params$distance_z_threshold)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 6  Genomic distribution

```{r plot_genomic, fig.cap="Genomic distribution of recombination rate and selection candidates."}
plots_genome <- lapply(sp_labels, function(sp) {
  df <- assigned[[sp]]$combined

  ggplot(df, aes(x = pos / 1e6, y = rec_rate_scaled)) +
    geom_point(aes(colour = factor(selection)), alpha = 0.4, size = 0.7) +
    geom_smooth(method = "loess", se = FALSE, colour = "grey30", linewidth = 0.8) +
    facet_wrap(~chr, scales = "free_x") +
    scale_colour_manual(
      values = c("0" = "#AAAAAA", "1" = "#C44E52"),
      labels = c("0" = "Background", "1" = "Candidate")
    ) +
    labs(title  = sp,
         x      = "Position (Mb)",
         y      = paste("Rec rate (", rec_unit_label, ")"),
         colour = NULL) +
    theme_classic(base_size = 9) +
    theme(strip.text = element_text(size = 8), legend.position = "bottom")
})

for (p in plots_genome) print(p)
```

---

## 7  Write output tables

```{r write_outputs}
# Combined assignment table (all species)
combined_all <- bind_rows(lapply(sp_labels, function(sp) {
  assigned[[sp]]$combined %>% mutate(species = sp)
}))

write_tsv(combined_all,
          file.path(params$output_dir, "gene_recombination_assignments.tsv"),
          na = "NA")

# GLM coefficient tables
write_tsv(coef_tbl,
          file.path(params$output_dir, "glm_coefficients.tsv"),
          na = "NA")

# Sensitivity results
if (nrow(sens_results) > 0)
  write_tsv(sens_results,
            file.path(params$output_dir, "sensitivity_analysis.tsv"),
            na = "NA")

cat("Output files written to:", params$output_dir, "\n")
cat("  gene_recombination_assignments.tsv\n")
cat("  glm_coefficients.tsv\n")
cat("  sensitivity_analysis.tsv\n")
```

---

## Session info

```{r session_info}
sessionInfo()
```
