---
title: "`r params$experiment_name`"
author: ""
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: yes
    toc_depth: 3
    code_folding: hide
    df_print: paged
params:
  experiment_name: "Gene Set MK Analysis (2DNS)"
  species_configs: !r list()
  min_genes: 8
  n_permutations: 1000
  length_strata_n: 5
  fisher_p_threshold: 0.05
  fdr_q_threshold: 0.05
  perm_p_threshold: 0.05
  seed: 123
  candidate_type: "bs_any"
  output_dir: "results/2dns"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 5
)
set.seed(params$seed)
```

```{r packages, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(kableExtra)
  library(patchwork)
})
```

```{r helpers, include=FALSE}

`%||%` <- function(a, b) if (!is.null(a)) a else b

# ── Statistical helpers ────────────────────────────────────────────────────────

safe_fisher <- function(mat, alternative = c("two.sided", "greater")) {
  alternative <- match.arg(alternative)
  if (any(!is.finite(mat)) || any(rowSums(mat) == 0) || any(colSums(mat) == 0))
    return(list(p.value = NA_real_, estimate = NA_real_))
  res <- tryCatch(fisher.test(mat, alternative = alternative), error = function(e) NULL)
  if (is.null(res)) return(list(p.value = NA_real_, estimate = NA_real_))
  list(p.value  = unname(res$p.value),
       estimate = if (!is.null(res$estimate)) unname(as.numeric(res$estimate)) else NA_real_)
}

mk_statistic <- function(PN, PS, DN, DS) {
  log((DN + 0.5) / (DS + 0.5)) - log((PN + 0.5) / (PS + 0.5))
}

make_length_breaks <- function(lengths, n_strata = 5L) {
  if (length(lengths) < 3 || all(is.na(lengths))) return(NULL)
  breaks <- unique(as.numeric(quantile(lengths,
    probs = seq(0, 1, length.out = n_strata + 1L), na.rm = TRUE, type = 7)))
  if (length(breaks) < 3L) NULL else breaks
}

assign_length_strata <- function(lengths, breaks) {
  if (is.null(breaks)) return(rep.int(1L, length(lengths)))
  as.integer(cut(lengths, breaks = breaks, include.lowest = TRUE, labels = FALSE))
}

# ── MKT data helpers ───────────────────────────────────────────────────────────

mk_filter_bs <- function(df) {
  df %>% filter(
    !is.na(num_pol_NS) & num_pol_NS >= 1,
    !is.na(num_pol_S)  & num_pol_S  >= 1,
    !is.na(Dn_branch_specific) & Dn_branch_specific >= 1,
    !is.na(Ds_branch_specific) & Ds_branch_specific >= 1,
    (num_pol_NS + num_pol_S) >= 5
  )
}

get_positive_flag <- function(df, type) {
  switch(type,
    bs_impMKT = df$pos_branch_specific_impMKT %in% TRUE,
    bs_mkt    = df$pos_branch_specific_MKT    %in% TRUE,
    bs_any    = (df$pos_branch_specific_impMKT %in% TRUE) | (df$pos_branch_specific_MKT %in% TRUE),
    any       = (df$pos_standard_MKT %in% TRUE) | (df$pos_standard_impMKT %in% TRUE) |
                (df$pos_branch_specific_MKT %in% TRUE) | (df$pos_branch_specific_impMKT %in% TRUE),
    stop("Unknown candidate_type: '", type, "'. Use one of: bs_any, bs_impMKT, bs_mkt, any")
  )
}

get_ortholog_col <- function(df) {
  # branch_specific_MKT_results.tsv uses 'ortholog'; some versions rename to 'aestivum_ortholog'
  for (col in c("aestivum_ortholog", "ortholog")) {
    if (col %in% names(df)) return(col)
  }
  NA_character_
}

load_mkt_data <- function(output_dir, candidate_type) {
  f <- file.path(output_dir, "branch_specific_MKT_results.tsv")
  if (!file.exists(f))
    stop("File not found: ", f, "\n  -> Run branch_specific_MKT/render_analysis.R first.")
  mkt <- read_tsv(f, show_col_types = FALSE) %>% mk_filter_bs()
  mkt %>% mutate(is_candidate = get_positive_flag(., candidate_type))
}

build_gene_pool <- function(mkt) {
  len_val <- if ("number_sites" %in% names(mkt)) {
    as.numeric(mkt$number_sites)
  } else {
    as.numeric(mkt$num_pol_NS + mkt$num_pol_S)
  }
  tibble(
    gene_id = mkt$gene_id,
    PN      = as.numeric(mkt$num_pol_NS),
    PS      = as.numeric(mkt$num_pol_S),
    DN      = as.numeric(mkt$Dn_branch_specific),
    DS      = as.numeric(mkt$Ds_branch_specific),
    length  = len_val
  ) %>%
    filter(if_all(c(PN, PS, DN, DS, length), is.finite), length > 0) %>%
    distinct(gene_id, .keep_all = TRUE)
}

# ── Analysis layers ────────────────────────────────────────────────────────────

run_fisher_analysis <- function(tbl, species, min_genes = 8L) {
  required <- c("pathway_name", "PN", "PS", "DN", "DS", "n_genes")
  miss <- setdiff(required, names(tbl))
  if (length(miss) > 0)
    stop("Input CSV for '", species, "' is missing columns: ", paste(miss, collapse = ", "))

  tbl %>%
    transmute(
      pathways   = pathway_name,
      PN = as.numeric(PN), PS = as.numeric(PS),
      DN = as.numeric(DN), DS = as.numeric(DS),
      n_genes    = as.numeric(n_genes),
      set_length = if ("set_length" %in% names(tbl)) as.numeric(set_length) else NA_real_
    ) %>%
    filter(!is.na(pathways), pathways != "", n_genes >= min_genes,
           if_all(c(PN, PS, DN, DS), is.finite)) %>%
    rowwise() %>%
    mutate(
      mat    = list(matrix(c(DN, DS, PN, PS), 2, byrow = TRUE)),
      res_2s = list(safe_fisher(mat, "two.sided")),
      res_gt = list(safe_fisher(mat, "greater"))
    ) %>%
    ungroup() %>%
    mutate(
      odds_ratio  = map_dbl(res_gt, "estimate"),
      p_two_sided = map_dbl(res_2s, "p.value"),
      p_greater   = map_dbl(res_gt, "p.value"),
      direction   = case_when(
        is.na(odds_ratio) ~ NA_character_,
        odds_ratio > 1    ~ "DN/DS > PN/PS",
        odds_ratio < 1    ~ "DN/DS < PN/PS",
        TRUE              ~ "DN/DS = PN/PS"
      ),
      species = species
    ) %>%
    select(species, pathways, n_genes, set_length, PN, PS, DN, DS,
           odds_ratio, direction, p_two_sided, p_greater) %>%
    arrange(p_two_sided)
}

# Permutation null: sample n_genes from the MKT gene pool, stratified by gene
# length (using overall pool length distribution, since gene-pathway membership
# is not available from branch_specific_MKT_results.tsv).
run_permutation <- function(obs_row, gene_pool, length_breaks, n_perm = 1000L) {
  n_genes  <- as.integer(obs_row$n_genes)
  obs_stat <- mk_statistic(obs_row$PN, obs_row$PS, obs_row$DN, obs_row$DS)

  na_row <- tibble(
    n_permutations   = n_perm,
    observed_stat    = obs_stat,
    perm_p_one_sided = NA_real_,
    perm_p_two_sided = NA_real_,
    perm_stat_median = NA_real_,
    perm_stat_q025   = NA_real_,
    perm_stat_q975   = NA_real_
  )

  if (!is.finite(obs_stat) || !is.finite(n_genes) || n_genes < 2 || nrow(gene_pool) == 0)
    return(na_row)

  pool <- gene_pool %>%
    mutate(length_stratum = assign_length_strata(length, length_breaks))

  strata_tbl <- pool %>%
    count(length_stratum) %>%
    mutate(target_n = pmax(1L, as.integer(round(n_genes * n / sum(n)))))

  perm_stats <- vapply(seq_len(n_perm), function(i) {
    sampled <- bind_rows(lapply(seq_len(nrow(strata_tbl)), function(j) {
      s   <- strata_tbl$length_stratum[j]
      n_s <- strata_tbl$target_n[j]
      cands <- pool %>% filter(length_stratum == s)
      if (nrow(cands) == 0) cands <- pool
      cands %>% slice_sample(n = n_s, replace = n_s > nrow(cands))
    }))
    mk_statistic(sum(sampled$PN), sum(sampled$PS), sum(sampled$DN), sum(sampled$DS))
  }, numeric(1))

  tibble(
    n_permutations   = n_perm,
    observed_stat    = obs_stat,
    perm_p_one_sided = (1 + sum(perm_stats >= obs_stat, na.rm = TRUE)) / (n_perm + 1),
    perm_p_two_sided = (1 + sum(abs(perm_stats) >= abs(obs_stat), na.rm = TRUE)) / (n_perm + 1),
    perm_stat_median = median(perm_stats, na.rm = TRUE),
    perm_stat_q025   = unname(quantile(perm_stats, 0.025, na.rm = TRUE)),
    perm_stat_q975   = unname(quantile(perm_stats, 0.975, na.rm = TRUE))
  )
}

# Jackknife: leave-one-gene-out Fisher test on pathway.
# Requires gene_level_data with columns: gene_id, pathway_name, PN, PS, DN, DS.
run_jackknife <- function(pathway_id, species_name, gene_level_data,
                          min_genes = 8L, p_threshold = 0.05) {
  if (is.null(gene_level_data)) return(NULL)
  req <- c("gene_id", "pathway_name", "PN", "PS", "DN", "DS")
  if (!all(req %in% names(gene_level_data))) return(NULL)

  pw_genes <- gene_level_data %>%
    filter(pathway_name == pathway_id) %>%
    transmute(
      gene_id = as.character(gene_id),
      PN = as.numeric(PN), PS = as.numeric(PS),
      DN = as.numeric(DN), DS = as.numeric(DS)
    ) %>%
    filter(!is.na(gene_id), gene_id != "") %>%
    distinct(gene_id, .keep_all = TRUE) %>%
    filter(if_all(c(PN, PS, DN, DS), is.finite))

  if (nrow(pw_genes) < min_genes) return(NULL)

  obs_p <- safe_fisher(
    matrix(c(sum(pw_genes$DN), sum(pw_genes$DS), sum(pw_genes$PN), sum(pw_genes$PS)),
           2, byrow = TRUE), "two.sided"
  )$p.value

  bind_rows(lapply(seq_len(nrow(pw_genes)), function(i) {
    red <- pw_genes[-i, ]
    if (nrow(red) < max(2L, min_genes - 1L)) return(NULL)
    jk_p <- safe_fisher(
      matrix(c(sum(red$DN), sum(red$DS), sum(red$PN), sum(red$PS)), 2, byrow = TRUE),
      "two.sided"
    )$p.value
    tibble(
      species           = species_name,
      pathway_id        = pathway_id,
      removed_gene_id   = pw_genes$gene_id[i],
      n_genes_retained  = nrow(red),
      obs_p_two_sided   = obs_p,
      jk_p_two_sided    = jk_p,
      p_change          = jk_p - obs_p,
      lost_significance = isTRUE(obs_p < p_threshold) & isTRUE(jk_p >= p_threshold)
    )
  })) %>% arrange(desc(p_change))
}

# Consistent kable style
styled_kable <- function(df, caption = NULL) {
  df %>%
    kable(caption = caption) %>%
    kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                  full_width = FALSE, position = "left")
}
```

```{r load_data, include=FALSE}
species_cfg   <- params$species_configs
species_names <- names(species_cfg)

if (length(species_cfg) == 0)
  stop("params$species_configs is empty. Fill in SPECIES_LIST in render_2dns.R.")

# Pathway-level Fisher inputs (input_2DNS_*.csv)
fisher_inputs <- setNames(lapply(species_names, function(sp) {
  read_csv(species_cfg[[sp]]$input_file, show_col_types = FALSE)
}), species_names)

# Gene-level MKT data for permutation background and candidates
mkt_data <- setNames(lapply(species_names, function(sp) {
  tryCatch(
    load_mkt_data(species_cfg[[sp]]$output_dir, params$candidate_type),
    error = function(e) { warning(sprintf("[%s] %s", sp, conditionMessage(e))); NULL }
  )
}), species_names)

gene_pools <- setNames(lapply(species_names, function(sp) {
  if (is.null(mkt_data[[sp]]))
    return(tibble(gene_id = character(), PN = numeric(), PS = numeric(),
                  DN = numeric(), DS = numeric(), length = numeric()))
  build_gene_pool(mkt_data[[sp]])
}), species_names)

length_breaks_list <- setNames(lapply(species_names, function(sp) {
  make_length_breaks(gene_pools[[sp]]$length, params$length_strata_n)
}), species_names)

# Optional gene-level pathway data for jackknife (user-provided)
gene_level_data <- setNames(lapply(species_names, function(sp) {
  glf <- species_cfg[[sp]]$gene_level_file
  if (is.null(glf) || !file.exists(glf)) return(NULL)
  read_csv(glf, show_col_types = FALSE)
}), species_names)

jackknife_available <- any(sapply(gene_level_data, Negate(is.null)))
```

---

# Overview

This report implements a multi-layer gene set MK test:

| Layer | Test | Input |
|-------|------|-------|
| 1 | Fisher exact (two-sided + one-sided) | Pathway-level PN/PS/DN/DS aggregates |
| 2 | Benjamini-Hochberg FDR | Layer 1 p-values |
| 3 | Stratified permutation | MKT gene pool (length-matched null) |
| 4 | Jackknife leave-one-out | Gene-level pathway data (optional) |

Candidate genes (Layer type: **`r params$candidate_type`**) are identified from
`branch_specific_MKT_results.tsv` in each species output directory.

Gene pool for the permutation null is built from all MKT-analyzable genes
(branch-specific filter: PN ≥ 1, PS ≥ 1, DN ≥ 1, DS ≥ 1, PN+PS ≥ 5).

```{r overview_table, echo=FALSE}
tibble(
  Species    = species_names,
  Label      = sapply(species_names, function(sp) species_cfg[[sp]]$label %||% sp),
  `Pathways` = sapply(species_names, function(sp) nrow(fisher_inputs[[sp]])),
  `Pool size`= sapply(species_names, function(sp) nrow(gene_pools[[sp]])),
  `MKT file` = sapply(species_names, function(sp)
    ifelse(is.null(mkt_data[[sp]]), "MISSING", "OK"))
) %>%
  styled_kable(caption = "Input summary")
```

---

# Layer 1: Fisher Exact Test

```{r fisher_run, include=FALSE}
results_fisher <- bind_rows(lapply(species_names, function(sp) {
  run_fisher_analysis(fisher_inputs[[sp]], sp, params$min_genes)
})) %>%
  group_by(species) %>%
  mutate(
    q_bh_two_sided = p.adjust(p_two_sided, method = "BH"),
    q_bh_greater   = p.adjust(p_greater,   method = "BH")
  ) %>%
  ungroup()

fisher_sig <- results_fisher %>%
  filter(p_two_sided < params$fisher_p_threshold |
         p_greater   < params$fisher_p_threshold)
```

## Summary

```{r fisher_summary}
results_fisher %>%
  group_by(species) %>%
  summarise(
    `Pathways tested`     = n(),
    `p < 0.05 (two-sided)` = sum(p_two_sided < params$fisher_p_threshold, na.rm = TRUE),
    `p < 0.05 (one-sided)` = sum(p_greater   < params$fisher_p_threshold, na.rm = TRUE),
    `q < 0.05 (BH, two-sided)` = sum(q_bh_two_sided < params$fdr_q_threshold, na.rm = TRUE),
    `Median p (two-sided)` = round(median(p_two_sided, na.rm = TRUE), 4),
    .groups = "drop"
  ) %>%
  rename(Species = species) %>%
  styled_kable(caption = "Layer 1 Fisher test summary")
```

## P-value distributions

```{r fisher_plots, fig.height=6}
p1 <- ggplot(results_fisher, aes(p_two_sided)) +
  geom_histogram(bins = 40, fill = "steelblue", alpha = 0.75) +
  facet_wrap(~species, scales = "free_y") +
  labs(x = "p-value (two-sided)", y = "Count",
       title = "Two-sided Fisher test p-values") +
  theme_minimal(base_size = 11)

p2 <- results_fisher %>%
  filter(!is.na(odds_ratio), is.finite(odds_ratio), odds_ratio > 0) %>%
  ggplot(aes(odds_ratio)) +
  geom_histogram(bins = 40, fill = "darkorange", alpha = 0.75) +
  facet_wrap(~species, scales = "free_y") +
  scale_x_log10() +
  labs(x = "Odds ratio (log scale)", y = "Count",
       title = "Fisher test odds ratio distributions") +
  theme_minimal(base_size = 11)

p1 / p2
```

## Significant pathways (raw p < `r params$fisher_p_threshold`)

```{r fisher_sig_table}
if (nrow(fisher_sig) == 0) {
  cat("No pathways significant at p <", params$fisher_p_threshold, "\n")
} else {
  fisher_sig %>%
    mutate(
      `DN/DS` = round(DN / DS, 3),
      `PN/PS` = round(PN / PS, 3),
      p_two_sided = formatC(p_two_sided, format = "e", digits = 2),
      p_greater   = formatC(p_greater,   format = "e", digits = 2)
    ) %>%
    select(Species = species, Pathway = pathways, `n genes` = n_genes,
           DN, DS, `DN/DS`, PN, PS, `PN/PS`, direction,
           `p (two-sided)` = p_two_sided, `p (one-sided)` = p_greater) %>%
    styled_kable(caption = paste("Pathways with raw Fisher p <", params$fisher_p_threshold))
}
```

---

# Layer 2: BH FDR Correction

```{r fdr_summary}
fdr_sig <- results_fisher %>%
  filter(q_bh_two_sided < params$fdr_q_threshold |
         q_bh_greater   < params$fdr_q_threshold)

results_fisher %>%
  group_by(species) %>%
  summarise(
    `Pathways tested`         = n(),
    `q < 0.05 (two-sided BH)` = sum(q_bh_two_sided < params$fdr_q_threshold, na.rm = TRUE),
    `q < 0.05 (one-sided BH)` = sum(q_bh_greater   < params$fdr_q_threshold, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(Species = species) %>%
  styled_kable(caption = paste0("Layer 2 — BH FDR (threshold: ", params$fdr_q_threshold, ")"))
```

```{r fdr_plot}
ggplot(results_fisher %>% filter(is.finite(p_two_sided)),
       aes(x = -log10(p_two_sided), y = -log10(q_bh_two_sided),
           colour = species, alpha = p_two_sided < params$fisher_p_threshold)) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = -log10(params$fdr_q_threshold), linetype = "dashed", colour = "red") +
  scale_alpha_manual(values = c(`FALSE` = 0.3, `TRUE` = 0.9), guide = "none") +
  facet_wrap(~species) +
  labs(x = "-log10(raw p)", y = "-log10(BH q)",
       title = "Raw p vs BH-corrected q (two-sided)",
       colour = "Species") +
  theme_minimal(base_size = 11) + theme(legend.position = "none")
```

---

# Layer 3: Stratified Permutation

For each pathway significant at Layer 1, `r params$n_permutations` gene sets of
the same size are drawn from the MKT-analyzable gene pool, stratified by gene
length (proxy: total coding sites). The empirical p-value is the fraction of
permuted MK statistics as extreme as the observed value (+1 continuity correction).

**Note:** Because gene-to-pathway membership is not encoded in
`branch_specific_MKT_results.tsv`, the length stratification uses the *overall*
pool distribution rather than the pathway's specific gene composition. This is a
conservative null model: it controls for gene-length bias without over-matching to
the observed pathway.

```{r perm_run, include=FALSE}
.perm_empty <- tibble(
  species = character(), pathways = character(), n_genes = integer(),
  set_length = numeric(), PN = numeric(), PS = numeric(), DN = numeric(), DS = numeric(),
  direction = character(), fisher_p_two_sided = numeric(), fisher_p_greater = numeric(),
  n_permutations = integer(), observed_stat = numeric(),
  perm_p_one_sided = numeric(), perm_p_two_sided = numeric(),
  perm_stat_median = numeric(), perm_stat_q025 = numeric(), perm_stat_q975 = numeric(),
  perm_q_two_sided = numeric(), perm_q_one_sided = numeric()
)

results_perm <- if (nrow(fisher_sig) == 0) {
  .perm_empty
} else {
  bind_rows(lapply(seq_len(nrow(fisher_sig)), function(i) {
    row <- fisher_sig[i, ]
    sp  <- row$species
    run_permutation(
      obs_row       = row,
      gene_pool     = gene_pools[[sp]],
      length_breaks = length_breaks_list[[sp]],
      n_perm        = params$n_permutations
    ) %>%
      mutate(
        species            = sp,
        pathways           = row$pathways,
        n_genes            = row$n_genes,
        set_length         = row$set_length,
        PN = row$PN, PS = row$PS, DN = row$DN, DS = row$DS,
        direction          = row$direction,
        fisher_p_two_sided = row$p_two_sided,
        fisher_p_greater   = row$p_greater
      )
  })) %>%
    mutate(
      perm_q_two_sided = p.adjust(perm_p_two_sided, method = "BH"),
      perm_q_one_sided = p.adjust(perm_p_one_sided, method = "BH")
    )
}

perm_sig <- results_perm %>%
  filter(perm_p_two_sided < params$perm_p_threshold |
         perm_p_one_sided < params$perm_p_threshold)
```

## Summary

```{r perm_summary}
if (nrow(results_perm) == 0) {
  cat("No pathways entered the permutation step (none significant at Layer 1).\n")
} else {
  results_perm %>%
    group_by(species) %>%
    summarise(
      `Pathways tested`       = n(),
      `perm p < 0.05 (two-s)` = sum(perm_p_two_sided < params$perm_p_threshold, na.rm = TRUE),
      `perm p < 0.05 (one-s)` = sum(perm_p_one_sided < params$perm_p_threshold, na.rm = TRUE),
      `Median perm p`         = round(median(perm_p_two_sided, na.rm = TRUE), 4),
      .groups = "drop"
    ) %>%
    rename(Species = species) %>%
    styled_kable(caption = "Layer 3 permutation summary")
}
```

## Fisher vs Permutation p-values

```{r perm_plot}
if (nrow(results_perm) > 0) {
  ggplot(results_perm %>% filter(is.finite(fisher_p_two_sided), is.finite(perm_p_two_sided)),
         aes(x = -log10(fisher_p_two_sided), y = -log10(perm_p_two_sided),
             colour = species, size = n_genes)) +
    geom_point(alpha = 0.7) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    facet_wrap(~species) +
    labs(x = "-log10(Fisher p, two-sided)",
         y = "-log10(Permutation p, two-sided)",
         title = "Fisher vs Permutation — two-sided p-values",
         colour = "Species", size = "n genes") +
    theme_minimal(base_size = 11) + theme(legend.position = "bottom")
}
```

## Significant pathways (permutation p < `r params$perm_p_threshold`)

```{r perm_sig_table}
if (nrow(perm_sig) == 0) {
  cat("No pathways significant after permutation at p <", params$perm_p_threshold, "\n")
} else {
  perm_sig %>%
    mutate(
      `DN/DS`            = round(DN / DS, 3),
      `PN/PS`            = round(PN / PS, 3),
      perm_p_two_sided   = formatC(perm_p_two_sided,   format = "e", digits = 2),
      perm_p_one_sided   = formatC(perm_p_one_sided,   format = "e", digits = 2),
      fisher_p_two_sided = formatC(fisher_p_two_sided, format = "e", digits = 2),
      set_length_kb      = if (!is.na(set_length[1])) round(set_length / 1000, 1) else NA_real_
    ) %>%
    select(Species = species, Pathway = pathways, `n genes` = n_genes,
           `Length (kb)` = set_length_kb,
           DN, DS, `DN/DS`, PN, PS, `PN/PS`, direction,
           `Fisher p` = fisher_p_two_sided,
           `Perm p (two-s)` = perm_p_two_sided,
           `Perm p (one-s)` = perm_p_one_sided) %>%
    styled_kable(caption = paste("Pathways with permutation p <", params$perm_p_threshold))
}
```

---

# Layer 4: Jackknife Sensitivity

```{r jackknife_header, results='asis'}
if (!jackknife_available) {
  cat(
    "> **Layer 4 skipped** — no gene-level pathway data provided.\n\n",
    "> To enable the jackknife, add `gene_level_file = \"path/to/file.csv\"` to each\n",
    "> species entry in `SPECIES_LIST` in `render_2dns.R`. The file must contain\n",
    "> columns: `gene_id`, `pathway_name`, `PN`, `PS`, `DN`, `DS`.\n\n"
  )
}
```

```{r jackknife_run, include=FALSE, eval=jackknife_available}
sig_pathways_for_jk <- perm_sig %>% distinct(species, pathways)

jackknife_results <- bind_rows(lapply(seq_len(nrow(sig_pathways_for_jk)), function(i) {
  sp <- sig_pathways_for_jk$species[i]
  pw <- sig_pathways_for_jk$pathways[i]
  run_jackknife(pw, sp, gene_level_data[[sp]],
                min_genes   = params$min_genes,
                p_threshold = params$perm_p_threshold)
}))

if (!is.null(jackknife_results) && nrow(jackknife_results) > 0) {
  jackknife_summary <- jackknife_results %>%
    group_by(species, pathway_id) %>%
    summarise(
      n_genes                  = n_genes_retained[1] + 1,
      obs_p_two_sided          = obs_p_two_sided[1],
      n_lost_significance      = sum(lost_significance, na.rm = TRUE),
      max_p_change             = max(p_change,          na.rm = TRUE),
      median_jk_p              = median(jk_p_two_sided, na.rm = TRUE),
      robust                   = n_lost_significance == 0,
      .groups = "drop"
    )
  outlier_genes <- jackknife_results %>%
    filter(lost_significance) %>%
    select(species, pathway = pathway_id, removed_gene_id, p_change)
} else {
  jackknife_summary <- tibble()
  outlier_genes     <- tibble()
}
```

```{r jackknife_output, eval=jackknife_available}
if (jackknife_available) {
  if (nrow(jackknife_summary) == 0) {
    cat("No pathways entered the jackknife (none significant at permutation step).\n")
  } else {
    jackknife_summary %>%
      mutate(
        obs_p_two_sided = formatC(obs_p_two_sided, format = "e", digits = 2),
        Robust          = ifelse(robust, "Yes", "No")
      ) %>%
      select(Species = species, Pathway = pathway_id, `n genes` = n_genes,
             `Obs p` = obs_p_two_sided, `n genes driving signal` = n_lost_significance,
             `Max p change` = max_p_change, Robust) %>%
      styled_kable(caption = "Layer 4 jackknife summary")

    if (nrow(outlier_genes) > 0) {
      cat("\n**Influential genes (removal causes loss of significance):**\n\n")
      outlier_genes %>%
        styled_kable(caption = "Genes driving the pathway signal")
    }
  }
}
```

---

# BS Candidate Genes

Candidate genes identified by `r params$candidate_type` from
`branch_specific_MKT_results.tsv` (MKT filter applied).

```{r candidates_load, include=FALSE}
candidates_all <- bind_rows(lapply(species_names, function(sp) {
  if (is.null(mkt_data[[sp]])) return(NULL)
  df    <- mkt_data[[sp]]
  orth  <- get_ortholog_col(df)
  tibble(
    species       = sp,
    label         = species_cfg[[sp]]$label %||% sp,
    gene_id       = df$gene_id,
    ortholog      = if (!is.na(orth)) df[[orth]] else NA_character_,
    is_candidate  = df$is_candidate,
    PN = df$num_pol_NS, PS = df$num_pol_S,
    DN = df$Dn_branch_specific, DS = df$Ds_branch_specific
  ) %>% filter(is_candidate)
}))
```

## Candidate gene counts

```{r candidates_summary}
bind_rows(lapply(species_names, function(sp) {
  if (is.null(mkt_data[[sp]]))
    return(tibble(Species = sp, Analyzable = NA, Candidates = NA, `With ortholog` = NA))
  df       <- mkt_data[[sp]]
  n_analyz <- nrow(df)
  n_cand   <- sum(df$is_candidate, na.rm = TRUE)
  orth     <- get_ortholog_col(df)
  n_orth   <- if (!is.na(orth)) sum(!is.na(df[[orth]][df$is_candidate]), na.rm = TRUE) else NA
  tibble(
    Species         = species_cfg[[sp]]$label %||% sp,
    Analyzable      = n_analyz,
    Candidates      = n_cand,
    `With ortholog` = n_orth,
    `% of analyzable` = if (n_analyz > 0) round(100 * n_cand / n_analyz, 1) else NA_real_
  )
})) %>%
  styled_kable(caption = paste("Candidate genes by species — type:", params$candidate_type))
```

## Candidate gene table

```{r candidates_table}
if (nrow(candidates_all) == 0) {
  cat("No candidate genes found.\n")
} else {
  candidates_all %>%
    mutate(
      `DN/DS` = round(DN / DS, 3),
      `PN/PS` = round(PN / PS, 3)
    ) %>%
    select(Species = label, `Gene ID` = gene_id, Ortholog = ortholog,
           PN, PS, DN, DS, `DN/DS`, `PN/PS`) %>%
    arrange(Species, `Gene ID`) %>%
    styled_kable(caption = "BS candidate genes with MKT counts")
}
```

> **Note:** Pathway membership of individual candidate genes cannot be determined
> from `branch_specific_MKT_results.tsv` alone. To assess which pathways contain
> candidate genes, provide a gene-level pathway annotation file via
> `gene_level_file` in `render_2dns.R`.

---

# Comprehensive Summary

```{r comprehensive_summary}
summary_all <- results_fisher %>%
  select(species, pathways, n_genes, DN, DS, PN, PS, p_two_sided, p_greater,
         q_bh_two_sided) %>%
  mutate(
    `L1 Fisher`  = p_two_sided   < params$fisher_p_threshold |
                   p_greater     < params$fisher_p_threshold,
    `L2 BH FDR` = q_bh_two_sided < params$fdr_q_threshold
  ) %>%
  left_join(
    results_perm %>% select(species, pathways, perm_p_two_sided, perm_p_one_sided),
    by = c("species", "pathways")
  ) %>%
  mutate(
    `L3 Permutation` = !is.na(perm_p_two_sided) &
      (perm_p_two_sided < params$perm_p_threshold |
       perm_p_one_sided < params$perm_p_threshold)
  )

# Add jackknife column if run and non-empty
.jk_avail <- jackknife_available &&
  exists("jackknife_summary") &&
  is.data.frame(jackknife_summary) &&
  nrow(jackknife_summary) > 0 &&
  "pathway_id" %in% names(jackknife_summary)

if (.jk_avail) {
  summary_all <- summary_all %>%
    left_join(jackknife_summary %>% select(species, pathways = pathway_id, robust),
              by = c("species", "pathways")) %>%
    rename(`L4 Robust` = robust)
} else {
  summary_all <- summary_all %>% mutate(`L4 Robust` = NA)
}

# Species-level counts
summary_all %>%
  group_by(species) %>%
  summarise(
    `Tested`          = n(),
    `L1 Fisher`       = sum(`L1 Fisher`,       na.rm = TRUE),
    `L2 BH FDR`       = sum(`L2 BH FDR`,       na.rm = TRUE),
    `L3 Permutation`  = sum(`L3 Permutation`,   na.rm = TRUE),
    `L4 Robust (JK)`  = sum(`L4 Robust` %in% TRUE),
    .groups = "drop"
  ) %>%
  rename(Species = species) %>%
  styled_kable(caption = "Pathways passing each layer per species")

# Pathways passing Fisher + Permutation
top_pathways <- summary_all %>%
  filter(`L1 Fisher`, `L3 Permutation`) %>%
  mutate(
    `DN/DS`  = round(DN / DS, 3),
    `PN/PS`  = round(PN / PS, 3),
    `Fisher p` = formatC(p_two_sided,   format = "e", digits = 2),
    `Perm p`   = formatC(perm_p_two_sided, format = "e", digits = 2)
  ) %>%
  select(Species = species, Pathway = pathways, `n genes` = n_genes,
         `DN/DS`, `PN/PS`, `Fisher p`, `Perm p`, `L2 BH FDR`, `L4 Robust`)

if (nrow(top_pathways) == 0) {
  cat("No pathways passed both Fisher and permutation filters.\n")
} else {
  top_pathways %>%
    styled_kable(caption = "Pathways passing Layer 1 (Fisher) and Layer 3 (Permutation)")
}
```

---

# Output files

```{r write_outputs}
dir.create(params$output_dir, recursive = TRUE, showWarnings = FALSE)

# Fisher results
write_tsv(results_fisher,
          file.path(params$output_dir, "fisher_results.tsv"),      na = "NA")

# Permutation results
if (nrow(results_perm) > 0)
  write_tsv(results_perm,
            file.path(params$output_dir, "permutation_results.tsv"), na = "NA")

# Comprehensive summary
write_tsv(summary_all,
          file.path(params$output_dir, "comprehensive_summary.tsv"), na = "NA")

# Candidates
if (nrow(candidates_all) > 0)
  write_tsv(candidates_all,
            file.path(params$output_dir, "candidates.tsv"),          na = "NA")

cat("TSV outputs written to:", params$output_dir, "\n",
    " - fisher_results.tsv\n",
    " - permutation_results.tsv\n",
    " - comprehensive_summary.tsv\n",
    " - candidates.tsv\n")
```

---

# Session info

```{r session_info}
sessionInfo()
```
