---
title: "McDonald-Kreitman Test — Standard and Imputed MKT"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: flatly
    fig_caption: true
params:
  merged_table: "step7_merged.tsv"
  output_table:  "step7_MKT_results.tsv"
  species: "focal species"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r libraries}
library(tidyverse)
library(kableExtra)
library(patchwork)
```

---

## 1  Load merged table

```{r load}
df <- read_tsv(
  params$merged_table,
  col_types = cols(.default = col_double(), gene_id = col_character()),
  na = c("NA", "NaN", "")
)

cat(sprintf("Loaded %d genes with %d columns\n", nrow(df), ncol(df)))
cat(sprintf("Species : %s\n", params$species))
cat(sprintf("Genes with Dn data : %d\n", sum(!is.na(df$Dn_counts))))
cat(sprintf("Genes with Ds data : %d\n", sum(!is.na(df$Ds_counts))))
```

---

## 2  Helper functions

```{r helpers}
# Fisher's exact test on the MK 2×2 table [[Pn, Dn], [Ps, Ds]]
mk_fisher <- function(Pn, Ps, Dn, Ds) {
  mapply(function(pn, ps, dn, ds) {
    if (any(is.na(c(pn, ps, dn, ds)))) return(NA_real_)
    if (any(c(pn, ps, dn, ds) < 0))   return(NA_real_)
    tryCatch(
      fisher.test(matrix(c(round(pn), round(ps), round(dn), round(ds)),
                         nrow = 2))$p.value,
      error = function(e) NA_real_
    )
  }, Pn, Ps, Dn, Ds, SIMPLIFY = TRUE)
}

# DoS = Dn/(Dn+Ds) − Pn/(Pn+Ps)  (Stoletzki & Eyre-Walker 2011)
compute_dos <- function(Pn, Ps, Dn, Ds) {
  div_frac  <- ifelse(!is.na(Dn) & !is.na(Ds) & (Dn + Ds) > 0,
                      Dn / (Dn + Ds), NA_real_)
  poly_frac <- ifelse(!is.na(Pn) & !is.na(Ps) & (Pn + Ps) > 0,
                      Pn / (Pn + Ps), NA_real_)
  div_frac - poly_frac
}

# α = 1 − (Ds × Pn) / (Dn × Ps)
compute_alpha <- function(Pn, Ps, Dn, Ds) {
  ifelse(!is.na(Dn) & !is.na(Ps) & Dn > 0 & Ps > 0,
         1 - (Ds * Pn) / (Dn * Ps), NA_real_)
}
```

---

## 3  Standard MKT

**Filter** (Stoletzki & Eyre-Walker 2011): Pn + Ps ≥ 5 and Pn, Ps, Dn, Ds ≥ 1.

```{r standard_mkt}
std <- df %>%
  mutate(
    Pn = num_pol_NS,
    Ps = num_pol_S,
    Dn = Dn_counts,
    Ds = Ds_counts,
    Standard_MKT        = compute_alpha(Pn, Ps, Dn, Ds),
    DoS_MKT_standard    = compute_dos(Pn, Ps, Dn, Ds),
    standard_MKT_pvalue = mk_fisher(Pn, Ps, Dn, Ds)
  )

# Build filter mask with explicit NA handling to avoid subscript-NA errors
std_mask <- with(std,
  !is.na(Pn) & Pn >= 1 &
  !is.na(Ps) & Ps >= 1 &
  !is.na(Dn) & Dn >= 1 &
  !is.na(Ds) & Ds >= 1 &
  (Pn + Ps) >= 5
)

std <- std %>% mutate(standard_MKT_p_adj = NA_real_)
if (any(std_mask)) {
  std$standard_MKT_p_adj[std_mask] <-
    p.adjust(std$standard_MKT_pvalue[std_mask], method = "BH")
}

std_filtered <- std %>% filter(std_mask)

cat(sprintf(
  "Standard MKT: %d / %d genes passed filter (Pn+Ps≥5, all cells≥1)\n",
  nrow(std_filtered), nrow(df)
))
cat(sprintf(
  "  Positively selected (DoS>0 & raw p<0.05): %d\n",
  sum(std_filtered$DoS_MKT_standard > 0 & std_filtered$standard_MKT_pvalue < 0.05,
      na.rm = TRUE)
))
```

---

## 4  Imputed MKT (impMKT)

Pn is corrected by subtracting the estimated SDM count (deleterious) before
applying the test. **Filter**: PnNeutral + Ps ≥ 5 and PnNeutral, Ps, Dn, Ds ≥ 1.

```{r impmkt}
imp <- std %>%
  mutate(
    # Re-cap PnNeutral_estimation to [0, Pn] for safety
    PnNeutral_safe         = pmax(0, pmin(PnNeutral_estimation, num_pol_NS)),
    standard_impMKT        = compute_alpha(PnNeutral_safe, Ps, Dn, Ds),
    DoS_impMKT_standard    = compute_dos(PnNeutral_safe, Ps, Dn, Ds),
    standard_impMKT_pvalue = mk_fisher(PnNeutral_safe, Ps, Dn, Ds)
  )

imp_mask <- with(imp,
  !is.na(PnNeutral_safe) & PnNeutral_safe >= 1 &
  !is.na(Ps) & Ps >= 1 &
  !is.na(Dn) & Dn >= 1 &
  !is.na(Ds) & Ds >= 1 &
  (PnNeutral_safe + Ps) >= 5
)

imp <- imp %>% mutate(standard_impMKT_p_adj = NA_real_)
if (any(imp_mask)) {
  imp$standard_impMKT_p_adj[imp_mask] <-
    p.adjust(imp$standard_impMKT_pvalue[imp_mask], method = "BH")
}

imp_filtered <- imp %>% filter(imp_mask)

cat(sprintf(
  "impMKT: %d / %d genes passed filter (PnNeutral+Ps≥5, all cells≥1)\n",
  nrow(imp_filtered), nrow(df)
))
cat(sprintf(
  "  Positively selected (DoS>0 & raw p<0.05): %d\n",
  sum(imp_filtered$DoS_impMKT_standard > 0 & imp_filtered$standard_impMKT_pvalue < 0.05,
      na.rm = TRUE)
))
```

---

## 5  Final output table

All columns from `merge_tables.py` output plus the 8 MKT statistics columns.

```{r final_table}
results <- imp %>%
  select(
    gene_id,
    number_sites, num_codons_eff,
    Pi_per_site, pi_S, pi_NS,
    thetaW_per_site, Tajimas_D,
    num_sites_S_egglib, num_sites_NS_egglib,
    polymorphism_count, num_pol_S, num_pol_NS,
    sample_size,
    PnMinus, PnGreater, PsMinus, PsGreater,
    ratioPs, PnNeutral_estimation,
    PnNeutral          = round(PnNeutral_safe),
    deleterious,
    Dn_counts, Ds_counts,
    Standard_MKT,        standard_impMKT,
    DoS_MKT_standard,    standard_MKT_pvalue,    standard_MKT_p_adj,
    DoS_impMKT_standard, standard_impMKT_pvalue, standard_impMKT_p_adj
  )

write_tsv(results, params$output_table, na = "NA")
cat(sprintf("Written: %s  (%d genes, %d columns)\n",
            params$output_table, nrow(results), ncol(results)))
```

---

## 6  Summary statistics

```{r summary_stats}
bind_rows(
  std_filtered %>% summarise(
    Test             = "Standard MKT",
    n_genes_tested   = n(),
    n_DoS_positive   = sum(DoS_MKT_standard   > 0,  na.rm = TRUE),
    n_sig_p05        = sum(standard_MKT_pvalue < 0.05, na.rm = TRUE),
    n_pos_selected   = sum(DoS_MKT_standard    > 0 & standard_MKT_pvalue < 0.05, na.rm = TRUE),
    pct_pos_selected = round(100 * n_pos_selected / n_genes_tested, 1),
    median_DoS       = round(median(DoS_MKT_standard, na.rm = TRUE), 4),
    median_alpha     = round(median(Standard_MKT,      na.rm = TRUE), 4)
  ),
  imp_filtered %>% summarise(
    Test             = "impMKT",
    n_genes_tested   = n(),
    n_DoS_positive   = sum(DoS_impMKT_standard   > 0,  na.rm = TRUE),
    n_sig_p05        = sum(standard_impMKT_pvalue < 0.05, na.rm = TRUE),
    n_pos_selected   = sum(DoS_impMKT_standard    > 0 & standard_impMKT_pvalue < 0.05, na.rm = TRUE),
    pct_pos_selected = round(100 * n_pos_selected / n_genes_tested, 1),
    median_DoS       = round(median(DoS_impMKT_standard, na.rm = TRUE), 4),
    median_alpha     = round(median(standard_impMKT,      na.rm = TRUE), 4)
  )
) %>%
  kbl(caption = sprintf("MKT summary — %s", params$species)) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```

---

## 7  Visualisations

```{r colours, include=FALSE}
CLR_STD  <- "#4C72B0"   # standard MKT (blue)
CLR_IMP  <- "#DD8452"   # imputed MKT (orange)
CLR_POS  <- "#C44E52"   # positive selection (red)
CLR_NEG  <- "#55A868"   # negative selection (green)
CLR_NTRL <- "#8C8C8C"   # not significant (grey)

theme_pub <- theme_classic(base_size = 11) +
  theme(
    plot.title    = element_text(size = 11, face = "bold"),
    plot.subtitle = element_text(size = 9,  colour = "grey40"),
    strip.text    = element_text(size = 10, face = "bold"),
    axis.title    = element_text(size = 10)
  )
theme_set(theme_pub)
```

### 7.1  Alignment statistics

Distribution of alignment length and sample size across genes.

```{r plot_alignment, fig.width=10, fig.height=4, fig.cap="Left: distribution of effective nucleotide sites per gene (number_sites). Right: distribution of the number of alleles analysed per gene (sample_size). Dashed line = median."}
p_len <- ggplot(df, aes(x = number_sites)) +
  geom_histogram(bins = 40, fill = CLR_STD, colour = "white") +
  geom_vline(xintercept = median(df$number_sites, na.rm = TRUE),
             linetype = "dashed", colour = "grey30") +
  labs(title = "Alignment length",
       subtitle = sprintf("median = %d nt", round(median(df$number_sites, na.rm = TRUE))),
       x = "Effective nucleotide sites", y = "Genes")

p_ss <- ggplot(df, aes(x = sample_size)) +
  geom_histogram(bins = 30, fill = CLR_STD, colour = "white") +
  geom_vline(xintercept = median(df$sample_size, na.rm = TRUE),
             linetype = "dashed", colour = "grey30") +
  labs(title = "Sample size",
       subtitle = sprintf("median = %d alleles", round(median(df$sample_size, na.rm = TRUE))),
       x = "Number of alleles", y = "Genes")

p_len + p_ss
```

### 7.2  Polymorphism statistics

```{r plot_pi, fig.width=12, fig.height=9, fig.cap="Population-genetics statistics. Top row: nucleotide diversity (pi) and Watterson theta per site. Bottom row: Tajima's D and synonymous vs non-synonymous diversity."}
p_pi <- ggplot(df, aes(x = Pi_per_site)) +
  geom_histogram(bins = 40, fill = CLR_STD, colour = "white") +
  geom_vline(xintercept = median(df$Pi_per_site, na.rm = TRUE),
             linetype = "dashed", colour = "grey30") +
  labs(title = expression(paste("Nucleotide diversity ", pi, " per site")),
       x = "Pi_per_site", y = "Genes")

p_tw <- ggplot(df, aes(x = thetaW_per_site)) +
  geom_histogram(bins = 40, fill = CLR_STD, colour = "white") +
  labs(title = expression(paste("Watterson's ", theta, " per site")),
       x = "thetaW_per_site", y = "Genes")

p_td <- ggplot(df, aes(x = Tajimas_D)) +
  geom_histogram(bins = 40, fill = CLR_STD, colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30") +
  labs(title = "Tajima's D", x = "Tajimas_D", y = "Genes")

p_piSNS <- df %>%
  filter(!is.na(pi_S) & !is.na(pi_NS)) %>%
  ggplot(aes(x = pi_S, y = pi_NS)) +
  geom_point(alpha = 0.3, size = 1, colour = CLR_STD) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey30") +
  labs(title = expression(paste(pi[NS], " vs ", pi[S])),
       subtitle = "Points above diagonal: excess NS diversity",
       x = expression(pi[S]), y = expression(pi[NS]))

(p_pi + p_tw) / (p_td + p_piSNS)
```

### 7.3  Polymorphism and divergence count distributions

```{r plot_counts, fig.width=12, fig.height=8, fig.cap="Distributions of synonymous (Ps) and non-synonymous (Pn) polymorphism counts (top row), and synonymous (Ds) and non-synonymous (Dn) divergence counts (bottom row) per gene."}
p_Ps <- ggplot(df, aes(x = num_pol_S)) +
  geom_histogram(bins = 30, fill = CLR_STD, colour = "white") +
  labs(title = "Synonymous polymorphisms (Ps)",
       x = "num_pol_S", y = "Genes")

p_Pn <- ggplot(df, aes(x = num_pol_NS)) +
  geom_histogram(bins = 30, fill = CLR_STD, colour = "white") +
  labs(title = "Non-synonymous polymorphisms (Pn)",
       x = "num_pol_NS", y = "Genes")

p_Ds <- ggplot(df %>% filter(!is.na(Ds_counts)), aes(x = Ds_counts)) +
  geom_histogram(bins = 30, fill = CLR_IMP, colour = "white") +
  labs(title = "Synonymous divergence (Ds)",
       x = "Ds_counts", y = "Genes")

p_Dn <- ggplot(df %>% filter(!is.na(Dn_counts)), aes(x = Dn_counts)) +
  geom_histogram(bins = 30, fill = CLR_IMP, colour = "white") +
  labs(title = "Non-synonymous divergence (Dn)",
       x = "Dn_counts", y = "Genes")

(p_Ps + p_Pn) / (p_Ds + p_Dn)
```

### 7.4  Polymorphism vs divergence scatter plots

```{r plot_scatter, fig.width=10, fig.height=4, fig.cap="Pn vs Ps per gene (left) and Dn vs Ds per gene (right). Dashed line = 1:1 ratio. Points above the diagonal have more NS than S counts."}
p_PnPs <- ggplot(df, aes(x = num_pol_S, y = num_pol_NS)) +
  geom_point(alpha = 0.25, size = 1, colour = CLR_STD) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey30") +
  labs(title = "Polymorphisms: Pn vs Ps",
       x = "Ps (num_pol_S)", y = "Pn (num_pol_NS)")

p_DnDs <- df %>%
  filter(!is.na(Dn_counts) & !is.na(Ds_counts)) %>%
  ggplot(aes(x = Ds_counts, y = Dn_counts)) +
  geom_point(alpha = 0.25, size = 1, colour = CLR_IMP) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey30") +
  labs(title = "Divergence: Dn vs Ds",
       x = "Ds (Ds_counts)", y = "Dn (Dn_counts)")

p_PnPs + p_DnDs
```

### 7.5  SDM estimation quality

```{r plot_sdm, fig.width=12, fig.height=9, fig.cap="SDM estimation diagnostics. Top-left: distribution of the neutral ratio ratioPs = Pn(>15%)/Ps(>15%). Dashed line at 1 = equal Pn/Ps. Top-right: distribution of the estimated SDM count (deleterious). Dashed line at 0. Bottom-left: corrected PnNeutral vs original Pn; points below the diagonal had SDMs removed. Bottom-right: fraction of Pn attributed to SDMs."}
p_ratio <- df %>%
  filter(!is.na(ratioPs) & is.finite(ratioPs)) %>%
  ggplot(aes(x = ratioPs)) +
  geom_histogram(bins = 40, fill = CLR_STD, colour = "white") +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey30") +
  labs(title = "Neutral ratio (ratioPs)",
       subtitle = expression(paste("P"[N]^{"(>15%)"}, " / P"[S]^{"(>15%)"})),
       x = "ratioPs", y = "Genes")

p_del <- df %>%
  filter(!is.na(deleterious)) %>%
  ggplot(aes(x = deleterious)) +
  geom_histogram(bins = 40, fill = CLR_STD, colour = "white") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30") +
  labs(title = "Estimated SDMs",
       subtitle = expression(paste("P"[SDM], " = P"[N]^{"(<15%)"},
                                   " − ratioPs × P"[S]^{"(<15%)"})),
       x = "deleterious", y = "Genes")

p_pnneut <- imp %>%
  filter(!is.na(num_pol_NS) & !is.na(PnNeutral_safe)) %>%
  ggplot(aes(x = num_pol_NS, y = PnNeutral_safe)) +
  geom_point(alpha = 0.25, size = 1, colour = CLR_STD) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey30") +
  labs(title = "PnNeutral vs Pn",
       subtitle = "Below diagonal = SDMs removed from Pn",
       x = "Pn (num_pol_NS)", y = "PnNeutral")

p_frac <- imp %>%
  filter(!is.na(num_pol_NS) & num_pol_NS > 0 & !is.na(deleterious)) %>%
  mutate(frac_del = pmax(0, deleterious) / num_pol_NS) %>%
  ggplot(aes(x = frac_del)) +
  geom_histogram(bins = 40, fill = CLR_STD, colour = "white") +
  labs(title = "Fraction of Pn attributed to SDMs",
       subtitle = "max(0, deleterious) / Pn",
       x = "Fraction SDM", y = "Genes")

(p_ratio + p_del) / (p_pnneut + p_frac)
```

### 7.6  MKT filter summary

```{r plot_filter, fig.width=7, fig.height=4, fig.cap="Number of genes at each analysis stage."}
filter_df <- tibble(
  Stage = factor(
    c("All genes",
      "Standard MKT\n(Pn+Ps≥5, cells≥1)",
      "impMKT\n(PnNeutral+Ps≥5, cells≥1)"),
    levels = c("All genes",
               "Standard MKT\n(Pn+Ps≥5, cells≥1)",
               "impMKT\n(PnNeutral+Ps≥5, cells≥1)")
  ),
  n = c(nrow(df), nrow(std_filtered), nrow(imp_filtered))
)

ggplot(filter_df, aes(x = Stage, y = n, fill = Stage)) +
  geom_col(width = 0.6, colour = "white") +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("grey70", CLR_STD, CLR_IMP)) +
  labs(title = "Genes passing MKT filters",
       x = NULL, y = "Number of genes") +
  expand_limits(y = max(filter_df$n) * 1.12) +
  theme(legend.position = "none")
```

### 7.7  Direction of Selection (DoS) distributions

```{r plot_dos, fig.width=10, fig.height=4, fig.cap="Distribution of DoS = Dn/(Dn+Ds) − Pn/(Pn+Ps) for genes passing each filter. DoS > 0 indicates positive selection, DoS < 0 indicates purifying selection."}
bind_rows(
  std_filtered %>% transmute(DoS = DoS_MKT_standard,    Test = "Standard MKT"),
  imp_filtered %>% transmute(DoS = DoS_impMKT_standard,  Test = "impMKT")
) %>%
  ggplot(aes(x = DoS, fill = Test)) +
  geom_histogram(binwidth = 0.05, colour = "white", alpha = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30") +
  facet_wrap(~Test, ncol = 2) +
  scale_fill_manual(values = c("Standard MKT" = CLR_STD, "impMKT" = CLR_IMP)) +
  labs(title = "Direction of Selection (DoS)",
       subtitle = "DoS = Dn/(Dn+Ds) − Pn/(Pn+Ps)",
       x = "DoS", y = "Genes") +
  theme(legend.position = "none")
```

### 7.8  α (proportion of adaptive substitutions)

```{r plot_alpha, fig.width=10, fig.height=4, fig.cap="Distribution of α = 1 − (Ds×Pn)/(Dn×Ps) for genes passing each filter."}
bind_rows(
  std_filtered %>% transmute(alpha = Standard_MKT,    Test = "Standard MKT"),
  imp_filtered %>% transmute(alpha = standard_impMKT, Test = "impMKT")
) %>%
  filter(is.finite(alpha)) %>%
  ggplot(aes(x = alpha, fill = Test)) +
  geom_histogram(bins = 40, colour = "white", alpha = 0.9) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey30") +
  facet_wrap(~Test, ncol = 2) +
  scale_fill_manual(values = c("Standard MKT" = CLR_STD, "impMKT" = CLR_IMP)) +
  labs(title = expression(paste(alpha, " — proportion of adaptive substitutions")),
       subtitle = expression(paste(alpha, " = 1 − (Ds×Pn)/(Dn×Ps)")),
       x = expression(alpha), y = "Genes") +
  theme(legend.position = "none")
```

### 7.9  p-value distributions

```{r plot_pval, fig.width=10, fig.height=4, fig.cap="Distribution of Fisher's exact test p-values for genes passing each filter. Enrichment near 0 indicates an excess of positive selection signals."}
bind_rows(
  std_filtered %>% transmute(pval = standard_MKT_pvalue,    Test = "Standard MKT"),
  imp_filtered %>% transmute(pval = standard_impMKT_pvalue, Test = "impMKT")
) %>%
  filter(!is.na(pval)) %>%
  ggplot(aes(x = pval, fill = Test)) +
  geom_histogram(bins = 20, colour = "white", alpha = 0.9) +
  geom_vline(xintercept = 0.05, linetype = "dashed", colour = "grey30") +
  facet_wrap(~Test, ncol = 2) +
  scale_fill_manual(values = c("Standard MKT" = CLR_STD, "impMKT" = CLR_IMP)) +
  labs(title = "Fisher's exact test p-value distribution",
       x = "p-value", y = "Genes") +
  theme(legend.position = "none")
```

### 7.10  Volcano plots (DoS vs −log₁₀ p)

```{r plot_volcano, fig.width=12, fig.height=5, fig.cap="Volcano plots. Red: DoS > 0 and p < 0.05 (positive selection). Green: DoS < 0 and p < 0.05 (purifying selection). Grey: not significant. Dashed lines at DoS = 0 and p = 0.05."}
bind_rows(
  std_filtered %>% transmute(
    DoS  = DoS_MKT_standard,
    pval = standard_MKT_pvalue,
    Test = "Standard MKT"
  ),
  imp_filtered %>% transmute(
    DoS  = DoS_impMKT_standard,
    pval = standard_impMKT_pvalue,
    Test = "impMKT"
  )
) %>%
  filter(!is.na(DoS) & !is.na(pval)) %>%
  mutate(
    log10p   = -log10(pval),
    category = case_when(
      DoS > 0 & pval < 0.05 ~ "Positive selection",
      DoS < 0 & pval < 0.05 ~ "Purifying selection",
      TRUE                  ~ "Not significant"
    )
  ) %>%
  ggplot(aes(x = DoS, y = log10p, colour = category)) +
  geom_point(alpha = 0.5, size = 1.0) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = 0,            linetype = "dashed", colour = "grey40") +
  facet_wrap(~Test, ncol = 2) +
  scale_colour_manual(values = c(
    "Positive selection"  = CLR_POS,
    "Purifying selection" = CLR_NEG,
    "Not significant"     = CLR_NTRL
  )) +
  labs(title = "Volcano plot — direction of selection",
       x = "DoS = Dn/(Dn+Ds) − Pn/(Pn+Ps)",
       y = expression(-log[10](p-value)),
       colour = NULL) +
  theme(legend.position = "bottom")
```

### 7.11  Classic McDonald-Kreitman plot

```{r plot_mk, fig.width=12, fig.height=5, fig.cap="McDonald-Kreitman plot. Points above the 1:1 dashed line have a higher dN/dS than pN/pS (consistent with positive selection). Red = DoS > 0 and p < 0.05."}
mk_std <- std_filtered %>%
  filter(Ps > 0 & Ds > 0) %>%
  transmute(
    pnps     = Pn / Ps,
    dnds     = Dn / Ds,
    selected = DoS_MKT_standard > 0 & standard_MKT_pvalue < 0.05,
    Test     = "Standard MKT"
  )

mk_imp <- imp_filtered %>%
  filter(Ps > 0 & Ds > 0) %>%
  transmute(
    pnps     = PnNeutral_safe / Ps,
    dnds     = Dn / Ds,
    selected = DoS_impMKT_standard > 0 & standard_impMKT_pvalue < 0.05,
    Test     = "impMKT"
  )

bind_rows(mk_std, mk_imp) %>%
  filter(is.finite(pnps) & is.finite(dnds)) %>%
  ggplot(aes(x = pnps, y = dnds, colour = selected)) +
  geom_point(alpha = 0.5, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(
    values = c("TRUE" = CLR_POS, "FALSE" = CLR_NTRL),
    labels = c("TRUE" = "DoS > 0 & p < 0.05", "FALSE" = "Non-significant"),
    name   = NULL
  ) +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~Test, ncol = 2) +
  labs(title = "McDonald-Kreitman plot",
       x = "Pn / Ps (polymorphism ratio, log scale)",
       y = "Dn / Ds (divergence ratio, log scale)") +
  theme(legend.position = "bottom")
```

---

## Session info

```{r session_info}
sessionInfo()
```
